---
title: Understanding the influence of precipitation timing and variability on dryland
  GPP
author: "Guy Lomax"
date: '2022-08-29'
output: html_notebook
---

This analysis uses Generalised Additive Models (GAMs) to explore the
relationship between different dimensions of precipitation variability and
vegetation productivity across African rangelands. The notebook calculates
mean values of different metrics of precipitation patterns and for gross primary
productivity for 2001-2019, along with a few other covariates, for each
rangeland pixel in Sub-Saharan Africa. A series of GAMs is then constructed to
identify relationships between these covariates and mean annual gross primary
productivity (GPP).

GPP data are from PML_V2 dataset (Gan et al., 2018; Zhang et al., 2019) at 8-day
temporal and 500m spatial resolution.
Precipitation data are from CHIRPS (Funk et al., 2015) at daily temporal
resolution and 0.05 degree spatial resolution.
All analysis conducted at 0.05 degree spatial resolution. Per-pixel metrics
are calculated as the mean value of 2001-2019 annual values.

```{r setup, include=FALSE}

options(repos = c(CRAN = "https://cran.rstudio.org"))
.libPaths("C:/Program Files/Microsoft/R Open/R-4.0.2/library")

# SET DIRECTORY TO SAVE FIGURES
plot_dir <- "figures/"

library(tidyverse)
library(terra)
library(sf)
library(tmap)
library(mgcv)
library(parallel)
library(gratia)
library(ggpubr)
library(patchwork)
library(beepr)

terraOptions(tempdir = "processed_data/temp/terra",
             memfrac = 0.8)

ggplot2::theme_set(theme_bw())

tmap_options(check.and.fix = TRUE)

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

nc <- 6 # number of cores
cl <- makeCluster(nc)
```

# Load and pre-process data

Raster data layers prepared, cropped and aggregated to 0.05 deg resolution
using Google Earth Engine.


```{r load_data, results = "hide", include = FALSE}

## EDIT FILE PATHS TO RELEVANT DIRECTORIES

# Africa country boundaries (Natural Earth dataset)
countries <- st_read("raw_data/vector/natural_earth/ne_110m_admin_0_countries_fixed.shp")
africa <- filter(countries, CONTINENT == "Africa")
africa_vct <- vect(africa)

# PML_V2 multi-annual mean GPP values 
# Convert GPP to raw units (scale by 0.01)
gpp_mean <- rast("raw_data/raster/PML_V2/pmlv2_gpp_mean.tif") * 0.01

# CHIRPS precipitation variables
map <- rast("raw_data/raster/CHIRPS/chirps_MAP.tif")
cv <- rast("raw_data/raster/CHIRPS/chirps_annual_CV.tif")
intensity <- rast("raw_data/raster/CHIRPS/chirps_intensity.tif")
dry <- rast("raw_data/raster/CHIRPS/chirps_dry_days.tif")
ugi <- rast("raw_data/raster/CHIRPS/chirps_UGi.tif")
pci <- rast("raw_data/raster/CHIRPS/chirps_PCI.tif")
pc_95 <- rast("raw_data/raster/CHIRPS/chirps_95pc.tif")
wet_95 <- rast("raw_data/raster/CHIRPS/chirps_wet_95pc.tif")
gt10 <- rast("raw_data/raster/CHIRPS/chirps_exceed_10.tif")
gt20 <- rast("raw_data/raster/CHIRPS/chirps_exceed_20.tif")
gt30 <- rast("raw_data/raster/CHIRPS/chirps_exceed_30.tif")

# CHIRPS season length and onset data 
season_length <- rast("processed_data/raster/seasonality/season_length.tif")
season_length_sd <- rast("processed_data/raster/seasonality/season_length_sd.tif")
season_onset_sd <- rast("processed_data/raster/seasonality/season_onset_sd.tif")

# ERA5-Land mean air temperature
meanT <- rast("raw_data/raster/ERA5-Land/era5_T.tif")

meanT_resample <- disagg(meanT, 2, method = "bilinear")

# MODIS mean daytime land surface temperature
lst <- rast("raw_data/raster/MODIS/modis_lst.tif")

# MODIS fire frequency 2001-2019
fire <- rast("raw_data/raster/MODIS/modis_burn_freq.tif") %>%
  classify(matrix(c(NA, 0), nrow = 1))

# iSDAsoil data (top 20cm)
soil <- rast("raw_data/raster/iSDAsoil/isda_0_20_cm.tif") %>%
  crop(map) # Fixing error in extent from export
names(soil) <- c("sand", "clay", "N", "P")

# Mask layers
# Aridity index
ai <- rast("raw_data/raster/WorldClim/aridity.tif")

# MODIS land cover classification IGBP
igbp <- rast("raw_data/raster/MODIS/IGBP_land_cover.tif")

# ESA WorldCover 10m land cover masks (2020)
# Additional higher-resolution layers excluding urban, cropland and forested land
esa_other <- rast("raw_data/raster/ESA/ESA_lc_frac.tif")
esa_tree <- rast("raw_data/raster/ESA/ESA_tree_frac.tif")

# Zero mask (remove pixels where more than 25% of sub-pixels are zero for all
# years in the dataset)
zero_mask <- rast("raw_data/raster/PML_V2/pmlv2_zero_mask.tif")

# CHIRPS seasonality ratio (for figures)
# sr > 1 indicates multiple rainy seasons per year
sr <- rast("raw_data/raster/chirps/seasonality_ratio.tif")

```


Clip and mask data layers:

```{r mask, include = FALSE}

igbp_index <- tibble(lc = c(6, 7, 8, 9, 10),
                     cover = factor(c("Closed shrublands",
                                      "Open shrublands",
                                      "Woody savannas",
                                      "Open savannas",
                                      "Grasslands")))

# Mask IGBP layers to African continental boundaries
igbp_mask <- mask(igbp, africa_vct)

# Most common assigned IGBP class

getMode <- function(x) {
keys <- na.omit(unique(x))
keys[which.max(tabulate(match(x, keys)))]
}

igbp_mode <- app(igbp_mask, getMode)

# # Write and read modal IGBP class layer (slow to calculate)
# writeRaster(igbp_mode, "processed_data/raster/modis/igbp_mode.tif",
#             overwrite = TRUE)
# 
# igbp_mode <- rast("processed_data/raster/modis/igbp_mode.tif")

# Aridity index mask (include areas with 0.05 < ai > 0.65)
ai_mask <- (ai > 500 & ai < 6500)

# ESA masks (exclude > 25% non-rangeland or > 50% tree cover)
esa_mask <- (esa_other < 0.25) & (esa_tree < 0.5)

# Combine masks
study_area_mask <- igbp_mode & ai_mask & esa_mask & zero_mask

# Study area mask function
sa_mask <- function(spatRaster) {
  masked <- spatRaster %>%
    mask(study_area_mask, maskvalues = c(0, NA))
  
  masked
}

# Create stack of covariate rasters and mask to study region

var_stack <- c(gpp_mean,
               igbp_mode, soil, fire, meanT_resample, lst,
               map, cv, intensity, dry, ugi, pc_95, wet_95,
               gt10, gt20, gt30,
               pci, season_length, season_length_sd, season_onset_sd, sr[[3]])

names(var_stack) <- c("gpp",
                      "lc",
                      "sand", "clay", "soilN", "soilP",
                      "fire", "meanT", "lst_mean",
                      "map", "cv", "intensity", "dry_days", "ugi",
                      "pc95", "wet95",
                      "gt10", "gt20", "gt30",
                      "pci", "season", "season_sd", "onset_sd", "sr")

var_stack_masked <- sa_mask(var_stack)

```

Convert whole raster to data frame

```{r to_df, cache = F, dependson = c("load", "mask"), include = FALSE}

df <- as.data.frame(var_stack_masked, xy = TRUE) %>%
  left_join(igbp_index)

```

# Exploratory Data Analysis

Plot histograms and scatter plots of key covariates.

Variable definitions:

|Variable name|Definition|Unit|Source dataset|
|-------------|----------|----|--------------|
|gpp|Gross Primary Productivity|gC m-2 yr-1|PML_V2|
|-------------|----------|----|--------------|
|map|Mean annual precipitation|mm yr-1|CHIRPS|
|cv|Coefficient of variation of MAP|%|CHIRPS|
|dry_days|Fraction of days per year with < 1mm precipitation|unitless 0-1|CHIRPS|
|intensity|Mean intensity of rainfall on a rainy day|mm day-1|CHIRPS|
|pc95|Fraction of daily rainfall above 95th percentile of all days|unitless 0-1|CHIRPS|
|wet95|Fraction of daily rainfall above 95th percentile of wet days|unitless 0-1|CHIRPS|
|gt10|Fraction of days per year with >10 mm rainfall|
|gt20|Fraction of days per year with >20 mm rainfall|
|gt30|Fraction of days per year with >30 mm rainfall|
|ugi|Unranked Gini index of daily precipitation|unitless 0-1|CHIRPS|
|pci|Precipitation concentration index|unitless 0-1|CHIRPS|
|-------------|----------|----|
|season|Average rainy season length|days|CHIRPS|
|season_sd|Standard deviation of rainy season length|days|CHIRPS|
|onset_sd|Standard deviation of rainy season onset|days|CHIRPS|
|-------------|----------|----|
|sand|Soil sand fraction 0-20cm|%|iSDAsoil|
|clay|Soil clay fraction 0-20cm|%|iSDAsoil|
|soilN|Soil nitrogen content 0-20cm|g/kg|iSDAsoil|
|soilP|Soil extractable phosphorus 0-20cm|ppm|iSDAsoil|
|meanT|Mean annual air temperature at 2m|degC|ERA5-Land|
|lst_mean|Mean daily land surface temperature|degC|MODIS Terra|
|fire|Mean fire frequency 2001-2019|yr-1|MODIS Terra|


```{r eda, eval = FALSE, echo = FALSE}

# Two-variable plots
# Sample to reduce display time
set.seed(123)
plot_sample <- sample_n(df, 4000) %>%
  filter(season > 0)    # Remove season length of zero (erroneous)

base_plot <- ggplot(plot_sample,
                    aes(fill = cover, colour = cover),
                    alpha = 0.6
                    ) +
  theme(axis.text = element_text(size = 8), axis.title = element_text(size = 10),
        legend.text = element_text(size = 10)) +
  labs(fill = "", colour = "") +
  theme_bw() +
  guides(colour = guide_legend(override.aes = list(size = 2)))

map_cv <- base_plot + geom_point(aes(x = map, y = cv), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = "CV (%)")
map_int <- base_plot + geom_point(aes(x = map, y = intensity), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = expression(SDII ~ (mm ~ d^-1)))
map_season <- base_plot + geom_point(aes(x = map, y = season), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = "Rainy season\nlength (d)")
map_onset_sd <- base_plot + geom_point(aes(x = map, y = onset_sd), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = expression(sigma[onset] ~ (d)))
map_ugi <- base_plot + geom_point(aes(x = map, y = ugi), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = "UGi")
map_wet95 <- base_plot + geom_point(aes(x = map, y = wet95), size = 0.6) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = expression(F["95w"]))

p1 <- ggarrange(map_cv, map_ugi, map_season, map_onset_sd, map_int, map_wet95,
          ncol = 2, nrow = 3, common.legend = TRUE)

gpp_expr <- expression(GPP ~ (gC ~ m^-2 ~ yr^-1))

gpp_sand <- base_plot + geom_point(aes(x = sand, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Soil sand fraction", y = gpp_expr)
gpp_clay <- base_plot + geom_point(aes(x = clay, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Soil clay fraction", y = gpp_expr)
gpp_T <- base_plot + geom_point(aes(x = meanT, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Mean air temperature\n(\u00B0C)", y = gpp_expr)
gpp_lst <- base_plot + geom_point(aes(x = lst_mean, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Mean daytime\nLST (\u00B0C)", y = gpp_expr)
gpp_fire <- base_plot + geom_point(aes(x = fire, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = expression(Fire ~ frequency ~ (yr^-1)), y = gpp_expr)

gpp_map <- base_plot + geom_point(aes(x = map, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = expression(MAP ~ (mm ~ yr^-1)), y = gpp_expr)
gpp_sd <- base_plot + geom_point(aes(x = cv * map * 0.01, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "SD of annual\nprecipitation (mm yr-1)", y = gpp_expr)
gpp_cv <- base_plot + geom_point(aes(x = cv, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "CV of annual\nprecipitation (%)", y = gpp_expr)
gpp_int <- base_plot + geom_point(aes(x = intensity, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = expression(Precipitation ~ intensity ~(mm ~ d^-1)), y = gpp_expr)
gpp_dry <- base_plot + geom_point(aes(x = dry_days, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Fraction of dry\ndays per year", y = gpp_expr)
gpp_ugi <- base_plot + geom_point(aes(x = ugi, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "UGi", y = gpp_expr)
gpp_wet95 <- base_plot + geom_point(aes(x = wet95, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = expression(F["95w"]), y = gpp_expr)
gpp_pci <- base_plot + geom_point(aes(x = pci, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "PCI", y = gpp_expr)
gpp_season <- base_plot + geom_point(aes(x = season, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Wet season\nlength (days)", y = gpp_expr)
gpp_season_sd <- base_plot + geom_point(aes(x = season_sd, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "SD of wet\nseason length (days)", y = gpp_expr)
gpp_onset_sd <- base_plot + geom_point(aes(x = onset_sd, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "SD of wet\nseason onset (days)", y = gpp_expr)
gpp_soilN <- base_plot + geom_point(aes(x = soilN, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = expression(Soil ~ nitrogen ~ (g ~ kg^-1)), gpp_expr)
gpp_soilP <- base_plot + geom_point(aes(x = soilP, y = gpp), size = 0.6, alpha = 0.75) +
  labs(x = "Soil extractable\nphosphorus (ppm)", y = gpp_expr)

p2 <- ggarrange(gpp_map, gpp_cv, gpp_ugi, gpp_season, gpp_onset_sd, gpp_int, gpp_wet95,
                nrow = 4, ncol = 2, common.legend = TRUE)

p3 <- ggarrange(gpp_soilN, gpp_sand, gpp_T, gpp_lst, gpp_fire,
                nrow = 3, ncol = 2, common.legend = TRUE)

# Save figures to directory

ggsave(paste0(plot_dir, "p1.jpg"),
       plot = p1,
       width = 16, height = 20, units = "cm", dpi = 300)
ggsave(paste0(plot_dir, "p2.jpg"),
       plot = p2,
       width = 16, height = 24, units = "cm", dpi = 300)
ggsave(paste0(plot_dir, "p3.jpg"),
       plot = p3, width = 16, height = 20, units = "cm", dpi = 300)

```


Exploratory maps of key covariates:


```{r eda_maps, eval = FALSE, echo = FALSE, results = "hide"}

# Create base map for all maps
africa_basemap <- tm_shape(africa) + tm_borders() +
  tm_layout(main.title.position = 0.1,
            legend.position = c("left", "bottom"), legend.text.size = 0.7,
            legend.title.size = 0.85,
            legend.bg.color = "white", legend.bg.alpha = 0,
            legend.height = -0.4, legend.width = 0.6) +
  tm_graticules(n.x = 6, n.y = 6, col = "grey", labels.size = 0.6)

# Create raster to mask excluded regions
outside_sa <- mask(study_area_mask, study_area_mask, maskvalues = c(1, NA))

sa_basemap <- tm_shape(outside_sa) + 
  tm_raster(col = "values", palette = "grey85", legend.show = FALSE)

# Study area map
sa_map <- tm_shape(var_stack_masked[["lc"]]) +
  tm_raster(style = "cat", palette = "Accent",
            labels = as.character(igbp_index$cover), title = "") +
  sa_basemap +
  africa_basemap

# GPP across region
gpp_map <- tm_shape(var_stack_masked[["gpp"]] %>% mask(africa_vct)) + 
  tm_raster(n = 9, palette = "YlGn", style = "cont", title = expression(GPP ~ (gC ~ m^-2 ~ yr^-1)),
            breaks = seq(0, 2500, 500)) +
  sa_basemap +
  africa_basemap +
  tm_layout(main.title = "a)")

# Precipitation maps

map_map <- tm_shape(var_stack_masked[["map"]]) +
  tm_raster(style = "cont", palette = "YlGnBu", title = expression("MAP" ~ (mm ~ yr^-1)),
            breaks = seq(0, 1500, 300)) +
  sa_basemap +
  africa_basemap +
  tm_layout(main.title = "b)")

cv_map <- tm_shape(var_stack_masked[["cv"]]) +
  tm_raster(style = "cont", n = 10, palette = "RdPu", title = "CV of annual\nprecipitation (%)",
            breaks = seq(0, 50, 10)) +
  sa_basemap +
  africa_basemap

intensity_map <- tm_shape(var_stack_masked[["intensity"]]) +
  tm_raster(style = "cont", palette = "PRGn", title = expression(SDII ~ (mm ~ day^-1)),
            breaks = seq(0, 20, 4)) +
  sa_basemap +
  africa_basemap

ugi_map <- tm_shape(var_stack_masked[["ugi"]]) +
  tm_raster(style = "cont", n = 10, palette = "-PRGn", title = "UGi of daily\nprecipitation",
            breaks = seq(0.1, 0.5, 0.1), midpoint = 0.3) +
  sa_basemap +
  africa_basemap

wet95_map <- tm_shape(var_stack_masked[["wet95"]]) +
  tm_raster(style = "cont", n = 10, palette = "OrRd", title = "Fraction of precipitation\nover 95th percentile",
            breaks = seq(0, 0.25, 0.05)) +
  sa_basemap +
  africa_basemap

# Seasonality maps

pci_map <- tm_shape(var_stack_masked[["pci"]]) +
  tm_raster(style = "fixed", n = 10, palette = "Purples", title = "Precipitation concentration index",
            breaks = seq(0, 40, 5), midpoint = 20) +
  sa_basemap +
  africa_basemap

season_map <- tm_shape(var_stack_masked[["season"]] %>% mask(var_stack_masked[["season"]] > 0, maskvalues = c(NA, 0))) +
  tm_raster(style = "cont", palette = "RdYlBu", title = "Rainy season\nlength (days)",
            breaks = seq(50, 250, 50), midpoint = 150) +
  sa_basemap +
  africa_basemap

onset_sd_map <- tm_shape(var_stack_masked[["onset_sd"]]) +
  tm_raster(style = "cont", palette = "-RdYlBu", title = expression(sigma[onset] ~ (days)),
            breaks = seq(0, 50, 10)) +
  sa_basemap +
  africa_basemap

# Soil property maps

soilN_map <- tm_shape(var_stack_masked[["soilN"]]) +
  tm_raster(style = "cont", palette = "YlGn", title = expression(soilN ~ (g ~ kg^-1)),
            breaks = seq(0, 2, 0.4)) +
  sa_basemap +
  africa_basemap

soilP_map <- tm_shape(var_stack_masked[["soilP"]]) +
  tm_raster(style = "cont", palette = "YlGn", title = expression(soilP ~ (g ~ kg^-1)),
            breaks = seq(0, 20, 4)) +
  sa_basemap +
  africa_basemap

sand_map <- tm_shape(var_stack_masked[["sand"]]) +
  tm_raster(style = "cont", palette = "-BrBG", title = "Soil sand %",
            breaks = seq(0, 100, 20)) +
  sa_basemap +
  africa_basemap

clay_map <- tm_shape(var_stack_masked[["clay"]]) +
  tm_raster(style = "cont", palette = "-BrBG", title = "Soil clay %",
            breaks = seq(0, 100, 20)) +
  sa_basemap +
  africa_basemap

# Temperature maps

T_map <- tm_shape(var_stack_masked[["meanT"]]) +
  tm_raster(style = "cont", n = 10, palette = "-RdYlBu", title = "MAT (\u00B0C)",
            breaks = seq(10, 35, 5)) +
  sa_basemap +
  africa_basemap +
  tm_layout(main.title = "c)")

lst_map <- tm_shape(var_stack_masked[["lst_mean"]]) +
  tm_raster(style = "cont", n = 10, palette = "-RdYlBu", title = "LST (\u00B0C)",
            breaks = seq(20, 45, 5)) +
  sa_basemap +
  africa_basemap +
  tm_layout(main.title = "d)")

# Fire frequency map

fire_map <- tm_shape(var_stack_masked[["fire"]]) +
  tm_raster(style = "cont", palette = "YlOrRd", title = expression(Fire ~ frequency ~ (yr^-1)),
            breaks = seq(0, 0.8, 0.2)) +
  sa_basemap +
  africa_basemap

gpp_map
map_map
cv_map
intensity_map
ugi_map
wet95_map
pci_map
season_map
onset_sd_map

sand_map
clay_map
soilN_map
soilP_map
T_map
lst_map

# Save map figure to directory

multi_plot <- tmap_arrange(gpp_map, map_map, T_map, lst_map)

tmap_save(multi_plot,
          paste0(plot_dir, "multi_map_plot.jpg"),
          width = 20, height = 18, units = "cm", dpi = 200)

```

Calculate correlation matrix for all variables in dataset:

```{r correlation}

cor_matrix = df %>%
  select(gpp,map,cv,intensity,dry_days,ugi,pci,pc95,wet95,gt10,gt30,
         season,onset_sd,sand,clay,soilN,soilP,meanT,lst_mean,fire) %>%
  cor(method = "spearman")

var_names <- c("GPP", "MAP", "CV", "SDII", "dry", "UGi", "PCI", ":F[95]",
               ":F['95w']", ":D['10mm']", ":D['30mm']", "season", ":sigma[onset]",
               "sand", "clay", "soilN", "soilP", "meanT", "LST", "fire")
colnames(cor_matrix) <- var_names
rownames(cor_matrix) <- var_names

# Save correlation plot figure to directory

jpeg(paste0(plot_dir, "cor_plot.jpg"),
     height = 50, width = 50, units = "cm", res = 300)

corrplot::corrplot(cor_matrix, method = "number", type = "upper",
                               tl.cex = 3, number.cex = 1.8, cl.cex = 2)

dev.off()

```

We remove covariates with > 0.7 Spearman's Rank correlation coefficient with
other covariates, leaving the following:

- MAP
- CV
- UGi
- season
- sigma_onset
- SDII
- F_95w
- soilN
- sand
- meanT
- LST
- fire


# GAMs

We use GAMs to explore the relationships and interactions between
rainfall dimensions, non-rainfall covariates and GPP.

The dependent variable (GPP) is non-negative and variance tends to increase with
the mean. A Gamma distribution with a log link may therefore be appropriate
for modelling these data.

We also include a smooth function of latitude and longitude s(x,y) to include
spatial relationships and spatially correlated errors not accounted for by other
covariates.

To prepare for GAM modelling, we select only covariates retained from correlation
analysis above.

```{r gam_prep, cache = F, dependson = c("load", "mask", "to_df")}

# Select model variables and scale some to more consistent units (0-1)
df_model <- df %>%
  dplyr::select(x, y, gpp, map, cv, ugi, season, onset_sd, intensity, wet95,
                soilN, sand, meanT, lst_mean, fire)%>%
  filter(season > 0) %>%    # Remove pixels with season length == 0 (errors)
  mutate(gpp = gpp * 0.001,
         sand = sand * 0.01,
         map = map * 0.001,
         cv = cv * 0.01)

# Write intermediate CSV for modelling

# write_csv(df_model, "processed_data/csv/gpp_precip/africa_sample_df_model.csv")
# df_model <- read_csv("processed_data/csv/gpp_precip/africa_sample_df_model.csv")

```

Non-spatial model including all covariates

```{r gam_1, cache = F, dependson = c("load", "mask", "to_df", "gam_prep")}

t1 <- Sys.time()
mod1 <- bam(gpp ~ s(map, k = 16, bs = "ts") + s(cv, k = 8, bs = "ts") +
              s(ugi, k = 8, bs = "ts") +
              s(season, k = 12, bs = "ts") + s(onset_sd, k = 8, bs = "ts") +
              s(wet95, k = 6, bs = "ts") + s(intensity, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") + s(sand, k = 6, bs = "ts") +
              s(lst_mean, k = 8, bs = "ts") +
              s(meanT, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            cluster = nc)
t2 <- Sys.time()
t2 - t1

# # Save/load model results
# saveRDS(mod1, "processed_data/models/mod1.rds")
# mod1 <- readRDS("processed_data/models/mod1.rds")

# Model 1 results
summary(mod1)
plot(mod1, shade = TRUE, seWithMean = TRUE, scheme = 2,
     shift = coef(mod1)[1], pages = 2)
gam.check(mod1)

# concurvity(mod1)
# (concurvity(mod1, full = FALSE)$estimate)

beep(3)

```
This model greatly improves the overall fit. A remarkable 84% of deviance is
explained by the model, and the structure of the residuals is slightly improved.
All variables included have a significant effect, with the possible exception of
F95w (extreme rainfall).

Pairwise concurvity remains low, suggesting the variables are sufficiently
independent.

Now I add the spatial term to account for the spatial structure in the model
residuals.

```{r gam_2, cache = F, dependson = c("load", "mask", "to_df", "gam_prep")}

# Spatial model - high smoothing (k = c(8,8) for spatial term)
Sys.time()
t1 <- Sys.time()
mod2_k8 <- bam(gpp ~ s(map, k = 16, bs = "ts") + s(cv, k = 8, bs = "ts") +
              s(ugi, k = 8, bs = "ts") +
              s(season, k = 12, bs = "ts") + s(onset_sd, k = 8, bs = "ts") +
              s(wet95, k = 6, bs = "ts") + s(intensity, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") + s(sand, k = 6, bs = "ts") +
              s(lst_mean, k = 8, bs = "ts") +
              s(meanT, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              te(x, y, k = c(8, 8), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            cluster = nc)
t2 <- Sys.time()
t2 - t1

beep(5)

# Spatial model - medium smoothing (k = c(12,12) for spatial term)
Sys.time()
t1 <- Sys.time()
mod2_k12 <- bam(gpp ~ s(map, k = 16, bs = "ts") + s(cv, k = 8, bs = "ts") +
              s(ugi, k = 8, bs = "ts") +
              s(season, k = 12, bs = "ts") + s(onset_sd, k = 8, bs = "ts") +
              s(wet95, k = 6, bs = "ts") + s(intensity, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") + s(sand, k = 6, bs = "ts") +
              s(lst_mean, k = 8, bs = "ts") +
              s(meanT, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              te(x, y, k = c(12, 12), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            cluster = nc)
t2 <- Sys.time()
t2 - t1

beep(5)

# Spatial model - low smoothing (k = c(16,16) for spatial term)

Sys.time()
t1 <- Sys.time()
mod2_k16 <- bam(gpp ~ s(map, k = 16, bs = "ts") + s(cv, k = 8, bs = "ts") +
              s(ugi, k = 8, bs = "ts") +
              s(season, k = 12, bs = "ts") + s(onset_sd, k = 8, bs = "ts") +
              s(wet95, k = 6, bs = "ts") + s(intensity, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") + s(sand, k = 6, bs = "ts") +
              s(lst_mean, k = 8, bs = "ts") +
              s(meanT, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              te(x, y, k = c(16, 16), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            cluster = nc)
t2 <- Sys.time()
t2 - t1

beep(3)

# # Save/load model results

# saveRDS(mod2_k8, "processed_data/models/mod2_k8.rds")
# saveRDS(mod2_k12, "processed_data/models/mod2_k12.rds")
# saveRDS(mod2_k16, "processed_data/models/mod2_k16.rds")
# mod2_k8 <- readRDS("processed_data/models/mod2_k8.rds")
# mod2_k12 <- readRDS("processed_data/models/mod2_k12.rds")
# mod2_k16 <- readRDS("processed_data/models/mod2_k16.rds")

# Model 2 results

plot(mod2_k8, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)
plot(mod2_k12, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)
plot(mod2_k16, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)

print("Spatial term k = 8")
gam.check(mod2_k8)
print("Spatial term k = 12")
gam.check(mod2_k12)
print("Spatial term k = 16")
gam.check(mod2_k16)


beep(3)

```

The spatial models have an improved fit, with 94-96% deviance explained and a
relatively symmetrical distribution of residuals (with some outliers).


# Plot GAM results

```{r term_plots}

# Set plot output directory

plot_dir <- "figures/"

theme_set(theme_classic())

## Main GAM partial effect plots

# Extract smooth shapes as data frame using gratia

mod1_smooths <- gratia::smooth_estimates(mod1, unconditional = TRUE) %>%
  mutate(Model = "Non-spatial",
         ci_low= est - se * 1.96,
         ci_high = est + se * 1.96,
         map = map * 1000)
mod2_k8_smooths <- gratia::smooth_estimates(mod2_k8, unconditional = TRUE) %>%
  mutate(Model = "Spatial - high smoothing",
         ci_low= est - se * 1.96,
         ci_high = est + se * 1.96,
         map = map * 1000) %>%
  select(-c(x,y))
mod2_k12_smooths <- gratia::smooth_estimates(mod2_k12, unconditional = TRUE) %>%
  mutate(Model = "Spatial - medium smoothing",
         ci_low= est - se * 1.96,
         ci_high = est + se * 1.96,
         map = map * 1000) %>%
  select(-c(x,y))
mod2_k16_smooths <- gratia::smooth_estimates(mod2_k16, unconditional = TRUE) %>%
  mutate(Model = "Spatial - low smoothing",
         ci_low= est - se * 1.96,
         ci_high = est + se * 1.96,
         map = map * 1000) %>%
  select(-c(x,y))

# Combine models into single data frame
mod_smooths <- bind_rows(mod1_smooths,
                         mod2_k8_smooths,
                         mod2_k12_smooths,
                         mod2_k16_smooths)

# Read in data labels lookup table

var_labels <- read_csv("csv/var_labels.csv")

var_list <- var_labels$vars

plots <- list(length = length(var_list))

# Generate plots

plots <- map(var_list, function(var) {
  
  # Labels for each facet window
  # Create expressions for those with superscript/subscripts
  x_label <- var_labels$label[var_labels$vars == var]
  if (var_labels$parse[var_labels$vars == var]) {
    x_label <- parse(text = x_label)
  }
  
  # Plot data density plot
  data_df <- df_model %>%
    select(x, y, all_of(var)) %>%
    rename(z = 3)
  
  density <- ggplot(data_df, aes(x = z)) +
    geom_density(colour = "grey70", fill = "grey90", alpha = 0.6) +
    theme_void()
  
  # Plot model smooths
  smooth_df <- mod_smooths %>%
    select(smooth, type, est, se, ci_low, ci_high, Model,
             all_of(var)) %>%
    mutate(Model = factor(Model, levels = c("Non-spatial",
                                            "Spatial - low smoothing",
                                            "Spatial - medium smoothing",
                                            "Spatial - high smoothing")),
           Model = ordered(Model)) %>%
    rename(var0 = 8) %>%
    drop_na()
  
  plot <- ggplot(smooth_df, aes(x = var0, y = est)) +
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high,
                    fill = Model), alpha = 0.7) +
    geom_line(aes(colour = Model, linetype = Model), alpha = 0.8, lwd = 0.6) +
    scale_colour_manual(values = c("#AAAAAA", "#bae4b3", "#31a354","#006d2c")) +
    scale_linetype_manual(values = c("longdash", "solid", "solid", "solid")) +
    scale_fill_manual(values = c("#DDDDDD", "#daffd3", "#51c384","#208d4c")) +
    guides(alpha = "none") +
    ylim(-1.2, 1) +
    labs(x = x_label,
         y = "Partial effect")
  
  # Arrange into small multiples with patchwork package 
  density + plot +
    plot_layout(ncol = 1, nrow = 2, widths = 4, heights = c(1, 4))
})

# Create main GAM figure
gam_plot <- ggarrange(plotlist = plots, ncol = 3, nrow = 4, common.legend = TRUE)

# Save spatial term map (CHOOSE DIRECTORY PATH)
ggsave(paste0(plot_dir, "main_gam_plot.jpg"),
       plot = gam_plot,
       width = 20, height = 30, units = "cm", dpi = 300)

# Create map of spatial term for low smoothing spatial model

mod2_spatial_smooth <- gratia::smooth_estimates(
  mod2_k16,
  n = 300,
  smooth = "te(x,y)",
  dist = 0.01) %>%
  mutate(ci_low= est - se * 1.96,
         ci_high = est + se * 1.96)

spatial_rast <- mod2_spatial_smooth %>%
  select(x, y, est) %>%
  as.data.frame() %>%
  rast(crs = "epsg:4326") %>%
  mask(africa_vct) %>%
  crop(var_stack_masked)

spatial_plot_mask <- study_area_mask %>%
  crop(spatial_rast)

africa_basemap <- tm_shape(africa) + tm_borders() +
  tm_layout(legend.position = c("left", "bottom"), legend.text.size = 0.8,
            legend.title.size = 1,
            legend.bg.color = "white", legend.bg.alpha = 0,
            legend.height = -0.4) +
  tm_graticules(n.x = 6, n.y = 6, col = "grey", labels.size = 0.6)

# Create additional mask raster to overlay on excluded regions
outside_sa <- mask(study_area_mask, study_area_mask, maskvalues = c(1, NA))

sa_basemap_2 <- tm_shape(outside_sa) +
  tm_raster(col = "values", palette = "grey75", legend.show = FALSE)

rast_plot <- tm_shape(spatial_rast) +
  tm_raster(palette = "RdYlBu", breaks = c(-2, -1, 0, 1, 2), style = "cont",
            title = "Partial effect\nof spatial term") +
  sa_basemap_2 +
  africa_basemap

# Save spatial term map
tmap_save(rast_plot,
          paste0(plot_dir, "spatial_term.jpg"),
          width = 16, height = 12, units = "cm", dpi = 300)


```


Model diagnostic and residual plots

```{r diag_plots}

# Model diagnostic plots

mod1_diag <- appraise(mod1, n_bins = 60, point_alpha = 0.05, line_col = "#888888")
ggsave(paste0(plot_dir, "mod1_diag_plot.jpg"),
       plot = mod1_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k8_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(paste0(plot_dir, "mod2_k8_diag_plot.jpg"),
       plot = mod2_k8_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k12_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(paste0(plot_dir, "mod2_k12_diag_plot.jpg"),
       plot = mod2_k12_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k16_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(paste0(plot_dir, "mod2_k16_diag_plot.jpg"),
       plot = mod2_k16_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

# Residual map
resid_raster <- df_model %>%
  mutate(resids = residuals(mod2_k16)) %>%
  select(x, y, resids) %>%
  rast(crs = "EPSG:4326")

resid_plot <- tm_shape(resid_raster) + 
  tm_raster(palette = "RdYlBu", style = "cont",
            breaks = c(-0.5, -0.25, 0, 0.25, 0.5),
            title = "Model residuals") +
  sa_basemap_2 +
  africa_basemap

tmap_save(resid_plot,
          paste0(plot_dir, "mod2_resid_map.jpg"),
          width = 16, height = 12, unit = "cm", dpi = 300)

```

