---
title: Disentangling the environmental drivers of gross primary productivity in African rangelands - Season length and anomaly calculations
author: Guy Lomax
date: 2024-05-14
output: html_notebook

---

This notebook processes data on wet season length, onset dates and cessation dates
and exports clean raster layers of mean wet season length, season length SD and
onset date SD for sub-Saharan Africa, as well as annual season length and mean
onset date anomalies.

```{r setup, message = FALSE}

library(tidyverse)
library(terra)
library(sf)
library(tmap)
library(here)

```

Import data:

```{r load_data}

# Hydrological year dates
hYear_start <- rast(here("data", "raw", "raster", "season_variables",
                           "hYearStartMean.tif"))

# Mean season dates - single and double seasons

mean_season_dates_s <- rast(here("data", "raw", "raster", "season_variables",
                               "meanSeasonDatesSingle.tif"))
mean_season_dates_d <- rast(here("data", "raw", "raster", "season_variables",
                               "meanSeasonDatesDouble.tif"))

# Seasonal rainfall annual dates and anomalies - single seasons
# Note that annual dates are expressed relative to the hydrological year start
annual_season_dates_s <- rast(here("data", "raw", "raster", "season_variables",
                            "annualSeasonDatesSingleHYear.tif"))

date_names_s <- c("onset_", "cessation_")
year_names_s <- as.character(rep(2001:2019, each = 2))

layer_names_s <- paste0(date_names_s, year_names_s)

names(annual_season_dates_s) <- layer_names_s

# Seasonal rainfall dates and anomalies - double seasons
annual_season_dates_d <- rast(here("data", "raw", "raster", "season_variables",
                            "annualSeasonDatesDoubleHYear.tif"))

date_names_d <- c("onset1_", "cessation1_", "onset2_", "cessation2_")
year_names_d <- as.character(rep(2001:2019, each = 4))

layer_names_d <- paste0(date_names_d, year_names_d)

names(annual_season_dates_d) <- layer_names_d

# Seasonality ratio

seasonality_ratio <- rast(here("data", "raw", "raster", "season_variables",
                               "seasonalityRatio.tif")) %>%
  subset("amplitude_ratio") %>%
  crop(annual_season_dates_s)  # Crop to extent (GEE export issues)

# Africa country boundaries (Natural Earth dataset)
africa <- st_read(here("data", "raw", "vector", "natural_earth",
                       "ne_110m_admin_0_countries_fixed.shp")) %>%
  filter(CONTINENT == "Africa") %>%
  vect()

```

Mask and process single and double season rasters into single raster objects

``` {r pre-process}

# Clip rasters to same extent (GEE export issue)
hYear_start <- crop(hYear_start, annual_season_dates_s)

# Function to mask to study area and separate single from double season pixels
mask_function <- function(rast, double = FALSE) {
  
  rast_cropped <- crop(rast, seasonality_ratio)
  rast_zero <- any(rast_cropped != 0)
  
  if(double
     ) {
    seasonality_mask <- seasonality_ratio >= 1
  } else {
    seasonality_mask <- seasonality_ratio < 1
  }

  rast_masked <- rast_cropped %>%
    mask(seasonality_mask, maskvalues = c(0, NA)) %>%
    mask(rast_zero, maskvalues = c(0,NA))
}

# Mask single season rasters
mean_season_dates_s <- mask_function(mean_season_dates_s, FALSE)
annual_season_dates_s <- mask_function(annual_season_dates_s, FALSE)

# Mask double season rasters
mean_season_dates_d <- mask_function(mean_season_dates_d, TRUE)
annual_season_dates_d <- mask_function(annual_season_dates_d, TRUE)

# Calculate mean start and end dates relative to hydrological year start

mean_season_dates_s_hYear <- (mean_season_dates_s + 365 - hYear_start) %% 365
mean_season_dates_d_hYear <- (mean_season_dates_d + 365 - hYear_start) %% 365

# Calculate anomalies in onset and cessation dates
season_dates_anomaly_s <- annual_season_dates_s - mean_season_dates_s_hYear
season_dates_anomaly_d <- annual_season_dates_d - mean_season_dates_d_hYear

# Calculate average season lengths and merge
mean_length_s <- mean_season_dates_s_hYear[[2]] - mean_season_dates_s_hYear[[1]]
mean_length_d <- (mean_season_dates_d_hYear[[c(2,4)]] - mean_season_dates_d_hYear[[c(1,3)]])

```

Explore data and identify issues/errors:


```{r eda}

# summary(seasonal_anomalies_s)
# summary(seasonal_anomalies_d)

# Single season onset dates
onset_anomaly_s <- season_dates_anomaly_s[[str_detect(names(season_dates_anomaly_s), "onset")]]

a <- sum(values(onset_anomaly_s < -60) | values(onset_anomaly_s > mean_length_s), na.rm = TRUE)
b <- sum(!is.na(values(onset_anomaly_s)))
print("Single season onset")
a/b

# Double season onset dates
onset_anomaly_d <- season_dates_anomaly_d[[str_detect(names(season_dates_anomaly_d), "onset")]]

c <- sum(values(onset_anomaly_d < -60) | values(onset_anomaly_d > mean_length_d), na.rm = TRUE)
d <- sum(!is.na(values(onset_anomaly_d)))
print("Double season onset")
c/d

# Single season cessation dates
cessation_anomaly_s <- season_dates_anomaly_s[[str_detect(names(season_dates_anomaly_s), "cessation")]]

# summary(cessation_anomaly)
e <- sum(values(cessation_anomaly_s + mean_length_s < 0) | values(cessation_anomaly_s > 60), na.rm = TRUE)
f <- sum(!is.na(values(cessation_anomaly_s)))
print("Single season cessation")
e/f

# Double season cessation dates
cessation_anomaly_d <- season_dates_anomaly_d[[str_detect(names(season_dates_anomaly_d), "cessation")]]

# summary(cessation_anomaly)
g <- sum(values(cessation_anomaly_d + mean_length_d < 0) | values(cessation_anomaly_d > 60), na.rm = TRUE)
h <- sum(!is.na(values(cessation_anomaly_d)))
print("Double season cessation")
g/h

# Identify pixels that have values outside the expected range

onset_error_pixels <- sum(onset_anomaly_s < -60 | onset_anomaly_s > 60)
cessation_error_pixels <- sum(cessation_anomaly_s < -60 | cessation_anomaly_s > 60)
onset_error_pixels_d <- sum(onset_anomaly_d < -60 | onset_anomaly_d > 60)
cessation_error_pixels_d <- sum(cessation_anomaly_d < -60 | cessation_anomaly_d > 60)

all_onset_errors <- merge(onset_error_pixels, onset_error_pixels_d)
all_cessation_errors <- merge(cessation_error_pixels, cessation_error_pixels_d)

plot(all_onset_errors)
lines(africa)
plot(all_cessation_errors)
lines(africa)

```

Remove pixels/seasons where onset/cessation dates have not been properly
identified:

```{r date_anomaly_pre_processing}

# Create a mask for removing seasons/years where:
# (i) estimated onset is < 60 days before mean onset
# (ii) estimated cessation is > 60 days after mean cessation
# (iii) estimated onset is after mean cessation
# (iv) estimated cessation is before mean onset

anomaly_mask_s <- (onset_anomaly_s >= -60 &
                          onset_anomaly_s < mean_length_s) & 
                       (cessation_anomaly_s + mean_length_s > 0 &
                          cessation_anomaly_s <= 60)

anomaly_mask_d <- (onset_anomaly_d >= -60 &
                          onset_anomaly_d <= mean_length_d) & 
                       (cessation_anomaly_d + mean_length_d > 0 &
                          cessation_anomaly_d <= 60)

# Fully mask pixels where more than 25% of years/seasons include NAs
lack_of_data_s <- 
  (sum(anomaly_mask_s) > nlyr(anomaly_mask_s) * 0.75)
lack_of_data_d <- 
  (sum(anomaly_mask_d) > nlyr(anomaly_mask_d) * 0.75)

lack_of_data <- merge(lack_of_data_s, lack_of_data_d)

# Apply masks

NA_mask <- function(rast, double = FALSE) {
  if(double) {
    anomaly_mask <- anomaly_mask_d
  } else {
    anomaly_mask <- anomaly_mask_s
  }
  
  rast %>%
    mask(anomaly_mask, maskvalues = c(0, NA)) %>%
    mask(lack_of_data, maskvalues = c(0, NA))
}

onset_anomaly_s_masked <- NA_mask(onset_anomaly_s)
cessation_anomaly_s_masked <- NA_mask(cessation_anomaly_s)
onset_anomaly_d_masked <- NA_mask(onset_anomaly_d, double = TRUE)
cessation_anomaly_d_masked <- NA_mask(cessation_anomaly_d, double = TRUE)

# Apply the same mask to the raw season onset/cessation layers

onset_dates_s <- annual_season_dates_s[[grepl("onset", names(annual_season_dates_s))]] %>%
  NA_mask()
cessation_dates_s <- annual_season_dates_s[[grepl("cessation", names(annual_season_dates_s))]] %>%
  NA_mask()

onset_dates_d <- annual_season_dates_d[[grepl("onset", names(annual_season_dates_d))]] %>%
  NA_mask(double = TRUE)
cessation_dates_d <- annual_season_dates_d[[grepl("cessation", names(annual_season_dates_d))]] %>%
  NA_mask(double = TRUE)

# summary(onset_anomaly_s_masked)
# summary(cessation_anomaly_s_masked)
# summary(onset_anomaly_d_masked)
# summary(cessation_anomaly_d_masked)

```

After masking, calculate key long-term metrics and combine single and double
into a single layer:

```{r seasonal_metrics}

# Overall mean season length
mean_length_d_combined = sum(mean_length_d, na.rm = T)
mean_length_all = merge(mean_length_s, mean_length_d_combined)

# Calculate total season lengths and merge
annual_length_s <- cessation_dates_s - onset_dates_s

# For double, sum individual season lengths for each year
season_length_d <- cessation_dates_d - onset_dates_d

annual_length_d <- season_length_d[[seq(1, nlyr(season_length_d)-1, 2)]] +
  season_length_d[[seq(2, nlyr(season_length_d), 2)]]

annual_length_all <- merge(annual_length_s, annual_length_d)

names(annual_length_all) <- paste0("seasonLength_", 2001:2019)

# Annual onset date anomaly

mean_onset_date_s <- mean_season_dates_s_hYear[["onset"]]
mean_onset_date_d1 <- mean_season_dates_d_hYear[["onset1"]]
mean_onset_date_d2 <- mean_season_dates_d_hYear[["onset2"]]

onset_dates_d1 <- onset_dates_d[[seq(1, nlyr(season_length_d)-1, 2)]]
onset_dates_d2 <- onset_dates_d[[seq(2, nlyr(season_length_d), 2)]]

onset_date_anomaly_s <- mean_onset_date_s - onset_dates_s
onset_date_anomaly_d1 <- mean_onset_date_d1 - onset_dates_d1
onset_date_anomaly_d2 <- mean_onset_date_d2 - onset_dates_d2

# Mask onset anomalies > 60 days

onset_anomaly_mask <- merge(
  (abs(onset_date_anomaly_s) <= 60),
  (abs(onset_date_anomaly_d1) <= 60 & abs(onset_date_anomaly_d2) <= 60)
)

onset_date_anomaly_d <- mean(onset_date_anomaly_d1, onset_date_anomaly_d2)

onset_anomaly_all <- onset_date_anomaly_s %>%
  merge(onset_date_anomaly_d) %>%
  mask(onset_anomaly_mask, maskvalues = c(0,NA))

# Standard deviation of onset, cessation and length
# Need to define a function to calculate SD that tolerates NA values while
# returning NA for fully masked pixels

season_sd <- function(x) {
  if(all(is.na(x))) {
    NA
  } else {
    sd(x, na.rm = TRUE)
  }
}

# Apply to relevant spatRasters
season_length_sd <- app(annual_length_all, fun = season_sd)

# Calculate onset sd
season_onset_sd_s <- app(onset_anomaly_s_masked, fun = season_sd)
season_onset_sd_d <- app(onset_anomaly_d_masked, fun = season_sd)
season_onset_sd <- merge(season_onset_sd_s, season_onset_sd_d)

```

Export data:

```{r export}

# Export as data layers for use in modelling
writeRaster(mean_length_all, here("data", "processed", "raster",
                                  "season_length_mean.tif"),
            overwrite = TRUE)

writeRaster(season_length_sd, here("data", "processed", "raster",
                                   "season_length_sd.tif"),
            overwrite = TRUE)
writeRaster(season_onset_sd, here("data", "processed", "raster",
                                   "season_onset_sd.tif"),
            overwrite = TRUE)

# Export annual season lengths and season start anomalies
writeRaster(annual_length_all, here("data", "processed", "raster",
                                    "season_length_annual.tif"),
            overwrite = TRUE)

writeRaster(onset_anomaly_all, here("data", "processed", "raster",
                                         "season_onset_anomaly.tif"),
            overwrite = TRUE)



```

