---
title: Disentangling the environmental drivers of gross primary productivity in African rangelands - annual analysis
author: "Guy Lomax"
date: '2023-10-09'
output: html_notebook
---

This analysis uses Generalised Additive Models (GAMs) to explore the
relationship between different dimensions of precipitation variability and
vegetation productivity across African rangelands. This notebook models annually
integrated gross primary productivity each year from 2001 to 2019 as a function of
precipitation, temperature and other metrics for each rangeland pixel in Sub-
Saharan Africa.

GPP data are from PML_V2 dataset (Gan et al., 2018; Zhang et al., 2019) at 8-day
temporal and 500m spatial resolution.
Precipitation data are from CHIRPS (Funk et al., 2015) at daily temporal
resolution and 0.05 degree spatial resolution.
All analysis conducted at 0.05 degree spatial resolution. 

```{r setup, include=FALSE}

# If running for first time, run renv::restore() to install required packages
# renv::restore()

# Data management

library(tidyverse)
library(terra)
library(sf)
library(here)

# Analysis

library(mgcv)
library(parallel)
library(tictoc)
library(beepr)

# Visualisation
library(tmap)
library(gratia)
library(ggpubr)
library(patchwork)

# Directory to save figures
plot_dir <- here("results", "figures")

ggplot2::theme_set(theme_bw())

# Establish number of cores for parallel computing
nc <- (detectCores() / 4)

```

# Load and pre-process data

Raster data layers prepared, cropped and aggregated to 0.05 deg resolution
using Google Earth Engine.


```{r load_data, results = "hide", include = FALSE}

# Africa country boundaries (Natural Earth dataset)
countries <- st_read(here("data", "raw", "vector", "natural_earth",
                          "ne_110m_admin_0_countries_fixed.shp"))
africa <- filter(countries, CONTINENT == "Africa")
africa_vct <- vect(africa)

# Load multi-annual data frame for modelling

df <- read_csv(here("data", "processed", "csv", "df_annual.csv")) %>%
  select(cell, x, y, gpp, starts_with("ppt"), starts_with("season"), everything())


```

# Exploratory Data Analysis

Plot histograms and scatter plots of key covariates.

Variable definitions:

|Variable name|Definition|Unit|Source dataset|
|-------------|----------|----|--------------|
|gpp|Gross Primary Productivity|gC m-2 yr-1|PML_V2|
|-------------|----------|----|--------------|
|map|Mean annual precipitation|mm yr-1|CHIRPS|
|cv|Coefficient of variation of MAP|%|CHIRPS|
|dry_days|Fraction of days per year with < 1mm precipitation|unitless 0-1|CHIRPS|
|intensity|Mean intensity of rainfall on a rainy day|mm day-1|CHIRPS|
|pc95|Fraction of daily rainfall above 95th percentile of all days|unitless 0-1|CHIRPS|
|wet95|Fraction of daily rainfall above 95th percentile of wet days|unitless 0-1|CHIRPS|
|gt10|Fraction of days per year with >10 mm rainfall|
|gt30|Fraction of days per year with >30 mm rainfall|
|ugi|Unranked Gini index of daily precipitation|unitless 0-1|CHIRPS|
|-------------|----------|----|
|season|Average rainy season length|days|CHIRPS|
|onset_sd|Standard deviation of rainy season onset|days|CHIRPS|
|-------------|----------|----|
|sand|Soil sand fraction 0-20cm|%|iSDAsoil|
|clay|Soil clay fraction 0-20cm|%|iSDAsoil|
|soilN|Soil nitrogen content 0-20cm|g/kg|iSDAsoil|
|soilP|Soil extractable phosphorus 0-20cm|ppm|iSDAsoil|
|meanT|Mean annual air temperature at 2m|degC|ERA5-Land|
|fire|Mean fire frequency 2001-2019|yr-1|MODIS Terra|


Calculate correlation matrix for all variables in dataset:

```{r correlation, eval = FALSE}

ppt_var_names <- c("pptTotal", "pptAnomaly", "pptMeanDayAnomaly", "pptIntensity",
               "pptGt95Pc", "pptUgi", "seasonLength", "seasonOnsetAnomaly", "pptMean")
non_ppt_var_names <- c("tMean", "fire", "sand", "clay", "soilN", "soilP")

cor_matrix <- df %>%
  dplyr::select(gpp, all_of(ppt_var_names), all_of(non_ppt_var_names)) %>%
  cor(method = "spearman", use="pairwise.complete.obs")

# var_names <- c("GPP", "MAP", "CV", "SDII", "dry", "UGi", "PCI", ":F[95]",
#                ":F['95w']", ":D['10mm']", ":D['30mm']", "season", ":sigma[onset]",
#                "sand", "clay", "soilN", "soilP", "tMean", "fire")
# colnames(cor_matrix) <- var_names
# rownames(cor_matrix) <- var_names

# Save correlation plot figure to directory

png(paste0(plot_dir, "/cor_plot_annual.png"),
     height = 50, width = 50, units = "cm", res = 300)

corrplot::corrplot(cor_matrix, method = "number", type = "upper",
                               tl.cex = 3, number.cex = 1.8, cl.cex = 2)

while (!is.null(dev.list()))  dev.off()

```

We remove covariates with > 0.7 Spearman's Rank correlation coefficient with
other covariates, leaving the following:

- Mean annual precipitation
- Precipitation anomaly
- Precipitaton mean day anomaly
- Precipitation intensity
- F95w
- UGi
- Season length
- Mean air temperature
- Soil sand fraction
- Soil N content
- Fire occurrence

# GAMs

We use GAMs to explore the relationships and interactions between
rainfall dimensions, non-rainfall covariates and GPP.

The dependent variable (GPP) is non-negative and variance tends to increase with
the mean. A Gamma distribution with a log link may therefore be appropriate
for modelling these data.

We also include a smooth function of latitude and longitude te(x,y) to include
spatial relationships and spatially correlated errors not accounted for by other
covariates, as well as a smooth function of time s(year)

To prepare for GAM modelling, we select only covariates retained from correlation
analysis above.

```{r gam_prep}

# Select variables equivalent to multi-annual analysis

target_vars <- c(
  "gpp",
  "pptTotal", "pptMean", "pptAnomaly",
  "pptIntensity", "pptGt95Pc", "pptUgi",
  "pptMeanDayAnomaly", "seasonLength", "seasonOnsetAnomaly",
  "tMean", "fire", "sand", "soilN"
)

df_model <- df %>%
  select(cell, x, y, year, all_of(target_vars)) %>%
  mutate(pptTotal = 0.001 * pptTotal,
         pptMean = 0.001 * pptMean) %>%
  filter(!is.na(seasonOnsetAnomaly))


# Subset of 10% of pixels to reduce model fitting issues
set.seed(123)
unique_cells <- unique(df_model$cell)
sample_cells <- sample(unique_cells, length(unique_cells) * 0.1)

df_model_sample <- filter(df_model, cell %in% sample_cells)

write_csv(df_model,
          here("data", "processed", "csv", "df_model_annual.csv"))
write_csv(df_model_sample,
          here("data", "processed", "csv", "df_model_annual_sample.csv"))


```


Simple model dividing precipitation into mean annual precipitation and annual
precipitation anomaly. We include the same spatial terms as in the multi-annual
model, and also include a simple smooth term of year in order to capture any
general time-related trends and potential temporal autocorrelation. In this
model, we don't include any interaction terms between covariates or between the
spatial and temporal parts of the model

```{r standard_models}

df_model <- read_csv(here("data", "processed", "csv", "df_model_annual.csv"))
df_model_sample <- read_csv(here("data", "processed", "csv", "df_model_annual_sample.csv"))

cl <- makeCluster(nc)

# Model with mean precipitation and annual anomaly

tic()
mod1 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              s(year, k = 8, bs = "ts"),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

write_rds(mod1, here("data", "processed", "models", "mod1_annual_sample.rds"))

tic()
mod2_k8 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(8, 8), bs = "ts") +
              s(year, k = 8, bs = "ts"),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

write_rds(mod2_k8, here("data", "processed", "models", "mod2_k8_annual_sample.rds"))

tic()
mod2_k12 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(12, 12), bs = "ts") +
              s(year, k = 8, bs = "ts"),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

write_rds(mod2_k12, here("data", "processed", "models", "mod2_k12_annual_sample.rds"))

tic()
mod2_k16 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(16, 16), bs = "ts") +
              s(year, k = 8, bs = "ts"),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl
)
toc()

write_rds(mod2_k16, here("data", "processed", "models", "mod2_k16_annual_sample.rds"))

stopCluster(cl)

```

```{r standard_models_viz}

mod1 <- read_rds(here("data", "processed", "models", "mod1_annual_sample.rds"))
mod2_k8 <- read_rds(here("data", "processed", "models", "mod2_k8_annual_sample.rds"))
mod2_k12 <- read_rds(here("data", "processed", "models", "mod2_k12_annual_sample.rds"))
mod2_k16 <- read_rds(here("data", "processed", "models", "mod2_k16_annual_sample.rds"))


mod1_diag <- appraise(mod1, ci_alpha = 0.5, point_alpha = 0.05, n_bins = 50)
mod2_k8_diag <- appraise(mod2_k8, ci_alpha = 0.5, point_alpha = 0.05, n_bins = 50)
mod2_k12_diag <- appraise(mod2_k12, ci_alpha = 0.5, point_alpha = 0.05, n_bins = 50)
mod2_k16_diag <- appraise(mod2_k16, ci_alpha = 0.5, point_alpha = 0.05, n_bins = 50)

mod1_diag
mod2_k8_diag 
mod2_k12_diag
mod2_k16_diag

ggsave(here("results", "figures", "mod1_diag_annual.png"), mod1_diag,
       width = 26.5, height = 22, units = "cm", dpi = 200)
ggsave(here("results", "figures", "mod2_k8_diag_annual.png"), mod2_k8_diag,
       width = 26.5, height = 22, units = "cm", dpi = 200)
ggsave(here("results", "figures", "mod2_k12_diag_annual.png"), mod2_k16_diag,
       width = 26.5, height = 22, units = "cm", dpi = 200)
ggsave(here("results", "figures", "mod2_k16_diag_annual.png"), mod2_k16_diag,
       width = 26.5, height = 22, units = "cm", dpi = 200)

```


```{r temporal_autocorrelation_check}

mod_residuals <- df_model_sample %>%
  mutate(mod1_resids = residuals(mod1, type = "response"),
         mod2_k8_resids = residuals(mod2_k8, type = "response"),
         mod2_k12_resids = residuals(mod2_k12, type = "response"),
         mod2_k16_resids = residuals(mod2_k16, type = "response"))

mod_ac <- mod_residuals %>%
  select(cell, x, y, year, gpp, pptMean, ends_with("resids")) %>%
  group_by(cell) %>%
  arrange(year, .by_group = TRUE) %>%
  reframe(x = mean(x),
          y = mean(y),
          gpp = mean(gpp),
          pptMean = mean(pptMean),
          lag = 0:5,
          across(ends_with("resids"),
                 function(res) {acf(res, plot = F, lag.max = 5)$acf}))

# Histogram and map of autocorrelation
histograms <- list(length = 5)

for (i in 1:5) {
  hist <- mod_ac %>%
    filter(lag == i) %>%
    ggplot(aes(x = mod2_k12_resids)) +
    geom_histogram(bins = 100, fill = "lightgreen", colour = "grey") +
    geom_vline(xintercept = 0, colour = "red")
  
  histograms[[i]] <- hist
}

mod_ac_rast <- mod_ac %>%
  select(x, y, mod2_k12_resids) %>%
  rast(crs = "EPSG:4326")

tm_shape(mod_ac_rast) + tm_raster(style = "cont", palette = "RdBu") +
  tm_shape(africa) + tm_borders()

tm_shape(mod_ac_rast > 0.45) + tm_raster()

```





``` {r interaction_mods}

cl <- makeCluster(nc)

tic()
mod3 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              # ti(pptMean, pptAnomaly, k = c(12,12), bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              # ti(pptAnomaly, pptMeanDayAnomaly, k = c(12,12), bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8) +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              ti(x, y, k = c(12, 12), bs = "ts") +
              ti(year, k = 8, bs = "ts") +
              ti(x, y, year, d = c(2,1), bs = c("ts", "ts"), k = c(6,6)),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

# # Save/load model results

write_rds(mod3, here("data", "processed", "models", "mod3_annual.rds"))


# With more interactions

tic()
mod3_int <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptAnomaly, k = 8, bs = "ts") +
              ti(pptMean, pptAnomaly, k = c(6,6), bs = "ts") +
              s(pptMeanDayAnomaly, k = 8, bs = "ts") +
              ti(pptAnomaly, pptMeanDayAnomaly, k = c(6,6), bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetAnomaly, k = 8) +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              ti(x, y, k = c(12, 12), bs = "ts") +
              ti(year, k = 8, bs = "ts") +
              ti(x, y, year, d = c(2,1), bs = c("ts", "ts"), k = c(6,6)),
            data = df_model_sample,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

# # Save/load model results

write_rds(mod3_int, here("data", "processed", "models", "mod3_annual_int.rds"))

```


# Plot GAM results

```{r term_plots}

df_model <- read_csv(here("data", "processed", "csv", "df_model_annual.csv"))
df_model_sample <- read_csv(here("data", "processed", "csv", "df_model_annual_sample.csv"))
study_area_mask <- rast(here("data", "processed", "raster", "study_area_mask.tif"))

# Set plot output directory

plot_dir <- here("results", "figures")

theme_set(theme_classic())

## Main GAM partial effect plots

# Extract smooth shapes as data frame using gratia

mod1_smooths <- gratia::smooth_estimates(mod1, unconditional = TRUE) %>%
  mutate(Model = "Non-spatial",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000)
mod2_k8_smooths <- gratia::smooth_estimates(mod2_k8, unconditional = TRUE) %>%
  mutate(Model = "Spatial - low\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000) %>%
  select(-c(x,y))
mod2_k12_smooths <- gratia::smooth_estimates(mod2_k12, unconditional = TRUE) %>%
  mutate(Model = "Spatial - medium\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000) %>%
  select(-c(x,y))
mod2_k16_smooths <- gratia::smooth_estimates(mod2_k16, unconditional = TRUE) %>%
  mutate(Model = "Spatial - high\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000) %>%
  select(-c(x,y))


# Combine models into single data frame
mod_smooths <- bind_rows(mod1_smooths,
                         mod2_k8_smooths,
                         mod2_k12_smooths,
                         mod2_k16_smooths) %>%
  filter(type == "TPRS (shrink)")

# Read in data labels lookup table

var_labels <- read_csv(here("data", "raw", "csv", "var_labels_annual.csv"))

var_list <- var_labels$vars

plots <- list(length = length(var_list))

# Generate plots

plots <- map(var_list, function(var) {
  
  # Labels for each facet window
  # Create expressions for those with superscript/subscripts
  x_label <- var_labels$label[var_labels$vars == var]
  if (var_labels$parse[var_labels$vars == var]) {
    x_label <- parse(text = x_label)
  }
  
  # Plot data density plot
  data_df <- df_model_sample %>%
    select(x, y, all_of(var)) %>%
    rename(z = 3)
  
  density <- ggplot(data_df, aes(x = z)) +
    geom_density(colour = "grey70", fill = "grey90", alpha = 0.6) +
    theme_void()
  
  # Plot model smooths
  smooth_df <- mod_smooths %>%
    select(smooth, type, est, se, ci_low, ci_high, Model,
             all_of(var)) %>%
    mutate(Model = factor(Model, levels = c("Non-spatial",
                                            "Spatial - low\ncomplexity",
                                            "Spatial - medium\ncomplexity",
                                            "Spatial - high\ncomplexity")),
           Model = ordered(Model)) %>%
    rename(var0 = 8) %>%
    drop_na()
  
  # Calculate bins
  diff <- smooth_df$var0[2] - smooth_df$var0[1]

  df_bins <- data_df %>%
    mutate(bin = cut_interval(z, length = diff)) %>%
    count(bin, .drop = FALSE)

  alpha_list <- pmin(1, (df_bins$n / (nrow(df_model) / 100))) %>%
    rep(4)
  smooth_df$alpha <- alpha_list[1:400]
  
  
  plot <- ggplot(smooth_df, aes(x = var0, y = est,
                                alpha = alpha
                                )) +
    geom_hline(yintercept = 1, colour = "grey80", linetype = "longdash") +
    # geom_ribbon(aes(ymin = ci_low, ymax = ci_high,
    # fill = Model)) +
    geom_line(aes(colour = Model), lwd = 0.6) +
    scale_colour_manual(values = c("#888888", "#41ab5d", "#238443","#004529")) +
    scale_linetype_manual(values = c("longdash", "solid", "solid", "solid")) +
    scale_alpha_continuous(range = c(0.3, 1)) +
    # scale_fill_manual(values = c("#DDDDDD", "#daffd3", "#51c384","#208d4c")) +
    guides(alpha = "none") +
    ylim(0, 2.5) +
    labs(x = x_label,
         y = "Partial effect")
  
  if(var_labels$min_0[var_labels$vars == var] == 1) {
    plot <- plot + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
  }
  
  if(var_labels$midpoint[var_labels$vars == var] != -1) {
    midpoint <- var_labels$midpoint[var_labels$vars == var]
    plot <- plot + geom_vline(xintercept = midpoint, colour = "grey80", linetype = "longdash")
  }
  
  # Arrange into small multiples with patchwork package 
  density + plot +
    plot_layout(ncol = 1, nrow = 2, widths = 4, heights = c(1, 4))
})

# Create main GAM figure
gam_plot <- ggarrange(plotlist = plots, ncol = 3, nrow = 4, common.legend = TRUE)

# Save spatial term map
ggsave(here(plot_dir, "annual_gam_plot_exp_alpha.jpg"),
       plot = gam_plot,
       width = 20, height = 30, units = "cm", dpi = 300)

# # Create map of spatial term for low smoothing spatial model
# 
# mod2_spatial_smooth <- gratia::smooth_estimates(
#   mod2_k16,
#   n = 300,
#   smooth = "te(x,y)",
#   dist = 0.01) %>%
#   mutate(ci_low= est - se * 1.96,
#          ci_high = est + se * 1.96)
# 
# spatial_rast <- mod2_spatial_smooth %>%
#   select(x, y, est) %>%
#   as.data.frame() %>%
#   rast(crs = "epsg:4326") %>%
#   mask(africa_vct) %>%
#   crop(var_stack_masked)
# 
# spatial_plot_mask <- study_area_mask %>%
#   crop(spatial_rast)
# 
# africa_basemap <- tm_shape(africa) + tm_borders() +
#   tm_layout(legend.position = c("left", "bottom"), legend.text.size = 0.8,
#             legend.title.size = 1,
#             legend.bg.color = "white", legend.bg.alpha = 0,
#             legend.height = -0.4) +
#   tm_graticules(n.x = 6, n.y = 6, col = "grey", labels.size = 0.6)
# 
# # Create additional mask raster to overlay on excluded regions
# outside_sa <- mask(study_area_mask, study_area_mask, maskvalues = c(1, NA))
# 
# sa_basemap_2 <- tm_shape(outside_sa) +
#   tm_raster(col = "values", palette = "grey75", legend.show = FALSE)
# 
# rast_plot <- tm_shape(spatial_rast) +
#   tm_raster(palette = "RdYlBu", breaks = c(-2, -1, 0, 1, 2), style = "cont",
#             title = "Partial effect\nof spatial term") +
#   sa_basemap_2 +
#   africa_basemap
# 
# # Save spatial term map
# tmap_save(rast_plot,
#           paste0(plot_dir, "spatial_term_annual.jpg"),
#           width = 16, height = 12, units = "cm", dpi = 300)


```


The results are very similar to the multi-annual, except for two aspects:
1. The annual precipitation anomaly is clearly important, although it seems
linear (on a log scale), which suggests a concave-up relationship overall.
2. The ranges of the covariates tend to be larger (because they are annual
values rather than means), such that the bulk of the data tends to be squeezed
into a very small range of the x-axis. This presumably means it's harder to
judge/detect any effect of these covariates. Do I need to centre and transform
variables to make them meaningful for the GAM?

The very similar results suggest that the model is still being driven by spatial
gradients in covariates rather than differences over time. Perhaps I need to
separately build a model that looks only at anomalies in GPP (fractional or
absolute) as a function of other anomalies? I could focus on the variables that
seemed to have substantive effects in the mean analysis for starters. The
challenge of GAMs is that it's very difficult to compare the magnitude of the
effect between the spatial model and the temporal model, since the functions
are composite polynomials and since the model involves a log link that
means a unit increase in a covariate doesn't have a simple interpretation in
terms of the response.

Model diagnostic and residual plots

```{r diag_plots}

# Model diagnostic plots

mod1_diag <- appraise(mod1, n_bins = 60, point_alpha = 0.05, line_col = "#888888")
ggsave(paste0(plot_dir, "mod1_annual_diag_plot.jpg"),
       plot = mod1_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(paste0(plot_dir, "mod2_annual_diag_plot.jpg"),
       plot = mod2_diag,
       width = 20, height = 16, units = "cm", dpi = 300)


# # Residual map
# resid_raster <- df_model %>%
#   mutate(resids = residuals(mod2_k16)) %>%
#   select(x, y, resids) %>%
#   rast(crs = "EPSG:4326")
# 
# resid_plot <- tm_shape(resid_raster) + 
#   tm_raster(palette = "RdYlBu", style = "cont",
#             breaks = c(-0.5, -0.25, 0, 0.25, 0.5),
#             title = "Model residuals") +
#   sa_basemap_2 +
#   africa_basemap
# 
# tmap_save(resid_plot,
#           paste0(plot_dir, "mod2_resid_map.jpg"),
#           width = 16, height = 12, unit = "cm", dpi = 300)

```



Repeat model for separate MAP bins:

```{r binned_models}

# library(future)
# library(furrr)

df_model_sample <- read_csv(here("data", "processed", "csv", "df_model_annual_sample.csv"))

df_model_bins <- df_model_sample %>%
  mutate(map_bin = cut_width(pptMean, width = 0.4, boundary = 0)) %>%
  group_by(map_bin) %>%
  nest() %>%
  ungroup()

# Cut to only first three bins
df_model_bins <- df_model_bins[1:3,]

fit_gam <- function(data, spatial = FALSE, spatial_k = 12) {
  
  if (spatial) {
    model <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
                   s(pptAnomaly, k = 8, bs = "ts") +
                   s(pptMeanDayAnomaly, k = 8, bs = "ts") +
                   s(pptIntensity, k = 8, bs = "ts") +
                   s(pptGt95Pc, k = 8, bs = "ts") +
                   s(pptUgi, k = 8, bs = "ts") +
                   s(seasonLength, k = 8, bs = "ts") +
                   s(seasonOnsetAnomaly, k = 8, bs = "ts") +
                   s(tMean, k = 8, bs = "ts") +
                   s(fire, k = 8, bs = "ts") +
                   s(sand, k = 8, bs = "ts") +
                   s(soilN, k = 8, bs = "ts") +
                   te(x, y, k = c(spatial_k, spatial_k), bs = "ts") +
                   s(year, k = 8, bs = "ts"),
                 data = data,
                 method = "REML",
                 family = Gamma(link = "log"),
                 select = TRUE,
                 samfrac = 0.1,
                 cluster = cl)
  } else {
    model <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
                   s(pptAnomaly, k = 8, bs = "ts") +
                   s(pptMeanDayAnomaly, k = 8, bs = "ts") +
                   s(pptIntensity, k = 8, bs = "ts") +
                   s(pptGt95Pc, k = 8, bs = "ts") +
                   s(pptUgi, k = 8, bs = "ts") +
                   s(seasonLength, k = 8, bs = "ts") +
                   s(seasonOnsetAnomaly, k = 8, bs = "ts") +
                   s(tMean, k = 8, bs = "ts") +
                   s(fire, k = 8, bs = "ts") +
                   s(sand, k = 8, bs = "ts") +
                   s(soilN, k = 8, bs = "ts") +
                   s(year, k = 8, bs = "ts"),
                 data = data,
                 method = "REML",
                 family = Gamma(link = "log"),
                 select = TRUE,
                 samfrac = 0.1,
                 cluster = cl)
  }
  
  model
}

cl <- makeCluster(nc)
# plan("multisession", workers = nc)

# tic()
# mod1_binned <- df_model_bins %>%
#   mutate(mod1 = map(data, fit_gam, spatial = FALSE)) %>%
#   select(-data)
# toc()
# write_rds(mod1_binned, here("data", "processed", "models", "mod1_binned_annual.rds"))
# 
# tic()
# mod2_k8_binned <- df_model_bins %>%
#   mutate(mod2_k8 = map(data, fit_gam, spatial = TRUE, spatial_k = 8)) %>%
#   select(-data)
# toc()
# write_rds(mod2_k8_binned, here("data", "processed", "models", "mod2_k8_binned_annual.rds"))

tic()
mod2_k12_binned <- df_model_bins %>%
  mutate(mod2_k12 = map(data, fit_gam, spatial = TRUE, spatial_k = 12)) %>%
  select(-data)
toc()
write_rds(mod2_k12_binned, here("data", "processed", "models", "mod2_k12_binned_annual.rds"))

tic()
mod2_k16_binned <- df_model_bins %>%
  mutate(mod2_k16 = map(data, fit_gam, spatial = TRUE, spatial_k = 16)) %>%
  select(-data)
toc()
write_rds(mod2_k16_binned, here("data", "processed", "models", "mod2_k16_binned_annual.rds"))

stopCluster(cl)

mod1_binned <- read_rds(here("data", "processed", "models", "mod1_binned_annual.rds"))
mod2_k8_binned <- read_rds(here("data", "processed", "models", "mod2_k8_binned_annual.rds"))
mod2_k12_binned <- read_rds(here("data", "processed", "models", "mod2_k12_binned_annual.rds"))
mod2_k16_binned <- read_rds(here("data", "processed", "models", "mod2_k16_binned_annual.rds"))

# Join
models_binned_joined <- df_model_bins %>%
  left_join(mod1_binned) %>%
  left_join(mod2_k8_binned) %>%
  left_join(mod2_k12_binned) %>%
  left_join(mod2_k16_binned)


write_rds(models_binned_joined,
          here("data", "processed", "models", "binned_models_annual.rds"))


```

Visualise binned models:

```{r binned_models_viz}

models_binned <- read_rds(here("data", "processed", "models", "binned_models_annual.rds"))

# plot_gam <- function(gam) {
#   plot(gam, pages = 3, seWithMean = T, scheme = 2)
# }

# Set plot output directory

plot_dir <- here("results", "figures")

theme_set(theme_classic())

## Binned GAM partial effect plots

for (i in 1:nrow(models_binned)) {
  
  # Extract smooth shapes as data frame using gratia
  
  mod1_smooths <- gratia::smooth_estimates(models_binned$mod1[[i]], unconditional = TRUE) %>%
    mutate(Model = "Non-spatial",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000)
  mod2_k8_smooths <- gratia::smooth_estimates(models_binned$mod2_k8[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - low\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000) %>%
    select(-c(x,y))
  mod2_k12_smooths <- gratia::smooth_estimates(models_binned$mod2_k12[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - medium\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000) %>%
    select(-c(x,y))
  mod2_k16_smooths <- gratia::smooth_estimates(models_binned$mod2_k16[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - high\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000) %>%
    select(-c(x,y))
  
  # Combine models into single data frame
  mod_smooths <- bind_rows(mod1_smooths,
                           mod2_k8_smooths,
                           mod2_k12_smooths,
                           mod2_k16_smooths) %>%
    filter(type == "TPRS (shrink)")
  
  # Read in data labels lookup table
  
  var_labels <- read_csv(here("data", "raw", "csv", "var_labels_annual.csv"))
  
  var_list <- var_labels$vars
  
  plots <- list(length = length(var_list))
  
  # Generate plots
  
  plots <- map(var_list, function(var) {
    
    # Labels for each facet window
    # Create expressions for those with superscript/subscripts
    x_label <- var_labels$label[var_labels$vars == var]
    if (var_labels$parse[var_labels$vars == var]) {
      x_label <- parse(text = x_label)
    }
  
    # Plot data density plot
    data_df <- models_binned$data[[i]] %>%
      select(x, y, all_of(var)) %>%
      rename(z = 3)
    
    density <- ggplot(data_df, aes(x = z)) +
      geom_density(colour = "grey70", fill = "grey90", alpha = 0.6) +
      theme_void()
    density
    
    # Plot model smooths
    smooth_df <- mod_smooths %>%
      select(smooth, type, est, se, ci_low, ci_high, Model,
               all_of(var)) %>%
      mutate(Model = factor(Model, levels = c("Non-spatial",
                                              "Spatial - low\ncomplexity",
                                              "Spatial - medium\ncomplexity",
                                              "Spatial - high\ncomplexity")),
             Model = ordered(Model)) %>%
      rename(var0 = 8) %>%
      drop_na()
    
    plot <- ggplot(smooth_df, aes(x = var0, y = est)) +
      geom_hline(yintercept = 1, colour = "grey80") +
      geom_ribbon(aes(ymin = ci_low, ymax = ci_high,
                      fill = Model), alpha = 0.7) +
      geom_line(aes(colour = Model, linetype = Model), alpha = 0.8, lwd = 0.6) +
      scale_colour_manual(values = c("#AAAAAA", "#bae4b3", "#31a354","#006d2c")) +
      scale_linetype_manual(values = c("longdash", "solid", "solid", "solid")) +
      scale_fill_manual(values = c("#DDDDDD", "#daffd3", "#51c384","#208d4c")) +
      guides(alpha = "none") +
      ylim(0, 2.5) +
      labs(x = x_label,
           y = "Partial effect")
    
      # if(var_labels$min_0[var_labels$vars == var] == 1) {
      #   plot <- plot + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
      #   density <- density + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
      # }
    
    # Arrange into small multiples with patchwork package 
    density + plot +
        plot_layout(ncol = 1, nrow = 2, widths = 4, heights = c(1, 4))
    })
    
    # Create main GAM figure
    gam_plot <- ggarrange(plotlist = plots, ncol = 3, nrow = 4, common.legend = TRUE)
    
    # Save main GAM figure
    ggsave(here(plot_dir, paste0("binned_gam_annual_plot_exp_", i, ".jpg")),
           plot = gam_plot,
           width = 20, height = 30, units = "cm", dpi = 300)
  
}

```
