---
title: Disentangling the environmental drivers of gross primary productivity in African rangelands - main analysis
author: "Guy Lomax"
date: '2023-05-31'
output: html_notebook
---

This analysis uses Generalised Additive Models (GAMs) to explore the
relationship between different dimensions of precipitation variability and
vegetation productivity across African rangelands. The notebook calculates
mean values of different metrics of precipitation patterns and for gross primary
productivity for 2001-2019, along with a few other covariates, for each
rangeland pixel in Sub-Saharan Africa. A series of GAMs is then constructed to
identify relationships between these covariates and mean annual gross primary
productivity (GPP).

GPP data are from PML_V2 dataset (Gan et al., 2018; Zhang et al., 2019) at 8-day
temporal and 500m spatial resolution.
Precipitation data are from CHIRPS (Funk et al., 2015) at daily temporal
resolution and 0.05 degree spatial resolution.
All analysis conducted at 0.05 degree spatial resolution. Per-pixel metrics
are calculated as the mean value of 2001-2019 annual values.

```{r setup, include=FALSE}

# If running for first time, run renv::restore() to install required packages
# renv::restore()

# Data management

library(tidyverse)
library(terra)
library(sf)
library(here)

# Analysis

library(mgcv)
library(parallel)
library(tictoc)

# Visualisation
library(tmap)
library(gratia)
library(ggpubr)
library(patchwork)

# Directory to save figures
plot_dir <- here("results", "figures")

ggplot2::theme_set(theme_bw())

tmap_options(check.and.fix = TRUE)

# Establish number of cores for parallel computing
nc <- (detectCores() / 4)

```

# Load and pre-process data

Raster data layers prepared, cropped and aggregated to 0.05 deg resolution
using Google Earth Engine.


```{r load_data, results = "hide", include = FALSE}

# Africa country boundaries (Natural Earth dataset)
countries <- st_read(here("data", "raw", "vector", "natural_earth",
                          "ne_110m_admin_0_countries_fixed.shp"))
africa <- filter(countries, CONTINENT == "Africa")
africa_vct <- vect(africa)

# Load multi-annual data frame for modelling

df <- read_csv(here("data", "processed", "csv", "df_multi_annual.csv")) %>%
  select(cell, x, y, gpp, starts_with("ppt"), starts_with("season"), everything())

```



# Exploratory Data Analysis

Plot histograms and scatter plots of key covariates.

Variable definitions:

|Variable name|Definition|Unit|Source dataset|
|-------------|----------|----|--------------|
|gpp|Gross Primary Productivity|gC m-2 yr-1|PML_V2|
|-------------|----------|----|--------------|
|map|Mean annual precipitation|mm yr-1|CHIRPS|
|cv|Coefficient of variation of MAP|%|CHIRPS|
|dry_days|Fraction of days per year with < 1mm precipitation|unitless 0-1|CHIRPS|
|intensity|Mean intensity of rainfall on a rainy day|mm day-1|CHIRPS|
|pc95|Fraction of daily rainfall above 95th percentile of all days|unitless 0-1|CHIRPS|
|wet95|Fraction of daily rainfall above 95th percentile of wet days|unitless 0-1|CHIRPS|
|gt10|Fraction of days per year with >10 mm rainfall|unitless 0-1|CHIRPS|
|gt20|Fraction of days per year with >20 mm rainfall|unitless 0-1|CHIRPS|
|gt30|Fraction of days per year with >30 mm rainfall|unitless 0-1|CHIRPS|
|mean_day_sd|Standard deviation in the mean date of rainfall|days|CHIRPS|
|ugi|Unranked Gini index of daily precipitation|unitless 0-1|CHIRPS|
|pci|Precipitation concentration index|unitless 0-1|CHIRPS|
|-------------|----------|----|
|season|Average rainy season length|days|CHIRPS|
|season_sd|Standard deviation of rainy season length|days|CHIRPS|
|onset_sd|Standard deviation of rainy season onset|days|CHIRPS|
|-------------|----------|----|
|sand|Soil sand fraction 0-20cm|%|iSDAsoil|
|clay|Soil clay fraction 0-20cm|%|iSDAsoil|
|soilN|Soil nitrogen content 0-20cm|g/kg|iSDAsoil|
|soilP|Soil extractable phosphorus 0-20cm|ppm|iSDAsoil|
|tMean|Mean annual air temperature at 2m|degC|ERA5-Land|
|fire|Mean fire frequency 2001-2019|yr-1|MODIS Terra|

Calculate correlation matrix for all variables in dataset:

```{r correlation, eval = FALSE}

cor_matrix = df %>%
  dplyr::select(-cell, -x, -y) %>%
  cor(method = "spearman", use="pairwise.complete.obs")

var_names <- c("GPP", "MAP", "dry", "intensity", ":F['95w']", ":D['10mm']",
               ":D['30mm']", "UGi", "CV", ":sigma[meanDay]", "season", ":sigma[onset]",
               "MAT", "fire","sand", "clay", "soilN", "soilP")
colnames(cor_matrix) <- var_names
rownames(cor_matrix) <- var_names

# Save correlation plot figure to directory

png(here(plot_dir, "cor_plot.png"),
     height = 50, width = 50, units = "cm", res = 300)

corrplot::corrplot(cor_matrix, method = "number", type = "upper",
                               tl.cex = 3, number.cex = 1.8, cl.cex = 2)

while (!is.null(dev.list()))  dev.off()

```

We remove covariates with > 0.7 Spearman's Rank correlation coefficient with
other covariates, leaving the following:

- MAP
- CV
- UGi
- season
- sigma_onset
- mean_day_sd
- SDII
- F_95w
- soilN
- sand
- tMean
- fire

# GAMs

We use GAMs to explore the relationships and interactions between
rainfall dimensions, non-rainfall covariates and GPP.

The dependent variable (GPP) is non-negative and variance tends to increase with
the mean. A Gamma distribution with a log link may therefore be appropriate
for modelling these data.

We also include a smooth function of latitude and longitude s(x,y) to include
spatial relationships and spatially correlated errors not accounted for by other
covariates.

To prepare for GAM modelling, we select only covariates retained from correlation
analysis above.

```{r gam_prep, include = FALSE}

# Select model variables and scale some to more consistent units (0-1)
df_model <- df %>%
  dplyr::select(x, y, gpp, pptMean, pptIntensity, pptGt95Pc, pptUgi, pptCV,
                pptMeanDaySD, seasonLength, seasonOnsetSD,
                tMean, fire, sand, soilN, ) %>%
  filter(seasonLength > 0 & !is.na(seasonOnsetSD)) %>%    # Remove pixels with season length == 0 (errors)
  mutate(gpp = gpp * 0.001,
         sand = sand * 0.01,
         pptMean = pptMean * 0.001)

# Write intermediate CSV for modelling

# write_csv(df_model,
#           here("data", "processed", "csv", "df_model_multi_annual.csv"))


```

Non-spatial model including all covariates

```{r non_spatial_model_fit, eval = FALSE}

df_model <- read_csv(here("data", "processed", "csv", "df_model_multi_annual.csv"))

cl <- makeCluster(nc)

tic()
mod1 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptCV, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(pptMeanDaySD, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetSD, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

stopCluster(cl)

# Save model results
write_rds(mod1, here("data", "processed", "models", "mod1.rds"))


```

```{r non_spatial_model_vis}

mod1 <- read_rds(here("data", "processed", "models", "mod1.rds"))

# Model 1 results
summary(mod1)
plot(mod1, shade = TRUE, seWithMean = TRUE, scheme = 2,
     shift = coef(mod1)[1], pages = 2)
gam.check(mod1)

# concurvity(mod1)
# (concurvity(mod1, full = FALSE)$estimate)

```


This model greatly improves the overall fit. A remarkable 84% of deviance is
explained by the model, and the structure of the residuals is slightly improved.
All variables included have a significant effect, with the possible exception of
F95w (extreme rainfall).

Pairwise concurvity remains low, suggesting the variables are sufficiently
independent.

Now I add the spatial term to account for the spatial structure in the model
residuals.

```{r spatial_model_fit, eval = FALSE}

df_model <- read_csv(here("data", "processed", "csv", "df_model_multi_annual.csv"))

cl <- makeCluster(nc)

# Spatial model - high smoothing (k = c(8,8) for spatial term)
tic()
mod2_k8 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptCV, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(pptMeanDaySD, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetSD, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(8, 8), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            samfrac = 0.1,
            select = TRUE,
            cluster = cl)
toc()

write_rds(mod2_k8, here("data", "processed", "models", "mod2_k8.rds"))

# Spatial model - medium smoothing (k = c(12,12) for spatial term)
tic()
mod2_k12 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptCV, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(pptMeanDaySD, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetSD, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(12, 12), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

write_rds(mod2_k12, here("data", "processed", "models", "mod2_k12.rds"))

# Spatial model - low smoothing (k = c(16,16) for spatial term)

tic()
mod2_k16 <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
              s(pptCV, k = 8, bs = "ts") +
              s(pptIntensity, k = 8, bs = "ts") +
              s(pptGt95Pc, k = 8, bs = "ts") +
              s(pptUgi, k = 8, bs = "ts") +
              s(pptMeanDaySD, k = 8, bs = "ts") +
              s(seasonLength, k = 8, bs = "ts") +
              s(seasonOnsetSD, k = 8, bs = "ts") +
              s(tMean, k = 8, bs = "ts") +
              s(fire, k = 8, bs = "ts") +
              s(sand, k = 8, bs = "ts") +
              s(soilN, k = 8, bs = "ts") +
              te(x, y, k = c(16, 16), bs = "ts"),
            data = df_model,
            method = "REML",
            family = Gamma(link = "log"),
            select = TRUE,
            samfrac = 0.1,
            cluster = cl)
toc()

write_rds(mod2_k16, here("data", "processed", "models", "mod2_k16.rds"))


```


```{r spatial_model_vis}

# Load model results

mod2_k8 <- read_rds(here("data", "processed", "models", "mod2_k8.rds"))
mod2_k12 <- read_rds(here("data", "processed", "models", "mod2_k12.rds"))
mod2_k16 <- read_rds(here("data", "processed", "models", "mod2_k16.rds"))

# Model 2 results

plot(mod2_k8, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)
plot(mod2_k12, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)
plot(mod2_k16, shade = TRUE, seWithMean = TRUE, scheme = 2,
     pages = 3)

print("Spatial term k = 8")
gam.check(mod2_k8)
print("Spatial term k = 12")
gam.check(mod2_k12)
print("Spatial term k = 16")
gam.check(mod2_k16)

```


The spatial models have an improved fit, with 94-96% deviance explained and a
relatively symmetrical distribution of residuals (with some outliers).


# Plot GAM results

``` {r term_plots}

df_model <- read_csv(here("data", "processed", "csv", "df_model_multi_annual.csv"))
study_area_mask <- rast(here("data", "processed", "raster", "study_area_mask.tif"))

# Set plot output directory

plot_dir <- here("results", "figures")

theme_set(theme_classic())

## Main GAM partial effect plots

# Extract smooth shapes as data frame using gratia

mod1_smooths <- gratia::smooth_estimates(mod1, unconditional = TRUE) %>%
  mutate(Model = "Non-spatial",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000,
         pptCV = pptCV * 100)      # CHANGE THIS BACK WHEN I RERUN MODELS!
mod2_k8_smooths <- gratia::smooth_estimates(mod2_k8, unconditional = TRUE) %>%
  mutate(Model = "Spatial - low\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000,
         pptCV = pptCV * 100) %>%
  select(-c(x,y))
mod2_k12_smooths <- gratia::smooth_estimates(mod2_k12, unconditional = TRUE) %>%
  mutate(Model = "Spatial - medium\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000,
         pptCV = pptCV * 100) %>%
  select(-c(x,y))
mod2_k16_smooths <- gratia::smooth_estimates(mod2_k16, unconditional = TRUE) %>%
  mutate(Model = "Spatial - high\ncomplexity",
         ci_low= exp(est - se * 1.96),
         ci_high = exp(est + se * 1.96),
         est = exp(est),
         pptMean = pptMean * 1000,
         pptCV = pptCV * 100) %>%
  select(-c(x,y))

# Combine models into single data frame
mod_smooths <- bind_rows(mod1_smooths,
                         mod2_k8_smooths,
                         mod2_k12_smooths,
                         mod2_k16_smooths) %>%
  filter(type == "TPRS (shrink)")

# Read in data labels lookup table

var_labels <- read_csv(here("data", "raw", "csv", "var_labels.csv"))

var_list <- var_labels$vars

plots <- list(length = length(var_list))

# Generate plots

plots <- map(var_list, function(var) {
  
  # Labels for each facet window
  # Create expressions for those with superscript/subscripts
  x_label <- var_labels$label[var_labels$vars == var]
  if (var_labels$parse[var_labels$vars == var]) {
    x_label <- parse(text = x_label)
  }

  # Plot data density plot
  data_df <- df_model %>%
    select(x, y, all_of(var)) %>%
    rename(z = 3)
  
  density <- ggplot(data_df, aes(x = z)) +
    geom_density(colour = "grey70", fill = "grey90", alpha = 0.6) +
    theme_void()
  density
  
  # Plot model smooths
  smooth_df <- mod_smooths %>%
    select(smooth, type, est, se, ci_low, ci_high, Model,
             all_of(var)) %>%
    mutate(Model = factor(Model, levels = c("Non-spatial",
                                            "Spatial - low\ncomplexity",
                                            "Spatial - medium\ncomplexity",
                                            "Spatial - high\ncomplexity")),
           Model = ordered(Model)) %>%
    rename(var0 = 8) %>%
    drop_na()
  
  plot <- ggplot(smooth_df, aes(x = var0, y = est)) +
    geom_hline(yintercept = 1, colour = "grey80") +
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high,
                    fill = Model), alpha = 0.7) +
    geom_line(aes(colour = Model, linetype = Model), alpha = 0.8, lwd = 0.6) +
    scale_colour_manual(values = c("#AAAAAA", "#bae4b3", "#31a354","#006d2c")) +
    scale_linetype_manual(values = c("longdash", "solid", "solid", "solid")) +
    scale_fill_manual(values = c("#DDDDDD", "#daffd3", "#51c384","#208d4c")) +
    guides(alpha = "none") +
    ylim(0, 2.5) +
    labs(x = x_label,
         y = "Partial effect")
  
  if(var_labels$min_0[var_labels$vars == var] == 1) {
    plot <- plot + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
  }
  
  # Arrange into small multiples with patchwork package 
  density + plot +
    plot_layout(ncol = 1, nrow = 2, widths = 4, heights = c(1, 4))
})

# Create main GAM figure
gam_plot <- ggarrange(plotlist = plots, ncol = 3, nrow = 4, common.legend = TRUE)

# Save main GAM figure
ggsave(here(plot_dir, "main_gam_plot_exp.jpg"),
       plot = gam_plot,
       width = 20, height = 30, units = "cm", dpi = 300)

# Create map of spatial term for high complexity spatial model

mod2_spatial_smooth <- gratia::smooth_estimates(
  mod2_k16,
  n = 300,
  smooth = "te(x,y)",
  dist = 0.01) %>%
  mutate(ci_low= est - se * 1.96,
         ci_high = est + se * 1.96)

spatial_rast <- mod2_spatial_smooth %>%
  select(x, y, est) %>%
  as.data.frame() %>%
  rast(crs = "epsg:4326") %>%
  mask(africa_vct) %>%
  crop(study_area_mask)

spatial_plot_mask <- study_area_mask %>%
  crop(spatial_rast)

africa_basemap <- tm_shape(africa) + tm_borders() +
  tm_layout(legend.position = c("left", "bottom"), legend.text.size = 0.8,
            legend.title.size = 1,
            legend.bg.color = "white", legend.bg.alpha = 0,
            legend.height = -0.4) +
  tm_graticules(n.x = 6, n.y = 6, col = "grey", labels.size = 0.6)

# Create additional mask raster to overlay on excluded regions
outside_sa <- mask(study_area_mask, study_area_mask, maskvalues = c(1, NA))

sa_basemap_2 <- tm_shape(outside_sa) +
  tm_raster(palette = "grey75", legend.show = FALSE)

rast_plot <- tm_shape(spatial_rast) +
  tm_raster(palette = "RdYlBu", breaks = c(-2, -1, 0, 1, 2), style = "cont",
            title = "Partial effect\nof spatial term") +
  sa_basemap_2 +
  africa_basemap

# Save spatial term map
tmap_save(rast_plot,
          here(plot_dir, "spatial_term.jpg"),
          width = 16, height = 12, units = "cm", dpi = 300)


```


Model diagnostic and residual plots

```{r diag_plots}

# Model diagnostic plots

mod1_diag <- appraise(mod1, n_bins = 60, point_alpha = 0.05, line_col = "#888888")
ggsave(here(plot_dir, "mod1_diag_plot.jpg"),
       plot = mod1_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k8_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(here(plot_dir, "mod2_k8_diag_plot.jpg"),
       plot = mod2_k8_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k12_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(here(plot_dir, "mod2_k12_diag_plot.jpg"),
       plot = mod2_k12_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

mod2_k16_diag <- appraise(mod2_k8, n_bins = 60, point_alpha = 0.05, line_col = "#117733")
ggsave(here(plot_dir, "mod2_k16_diag_plot.jpg"),
       plot = mod2_k16_diag,
       width = 20, height = 16, units = "cm", dpi = 300)

# Residual map
resid_raster <- df_model %>%
  mutate(resids = residuals(mod2_k16)) %>%
  select(x, y, resids) %>%
  rast(crs = "EPSG:4326")

resid_plot <- tm_shape(resid_raster) + 
  tm_raster(palette = "RdYlBu", style = "cont",
            breaks = c(-0.5, -0.25, 0, 0.25, 0.5),
            title = "Model residuals") +
  sa_basemap_2 +
  africa_basemap

tmap_save(resid_plot,
          here(plot_dir, "mod2_resid_map.jpg"),
          width = 16, height = 12, unit = "cm", dpi = 300)

```

Repeat model for separate MAP bins:
```{r binned_models}

library(future)
library(furrr)

df_model <- read_csv(here("data", "processed", "csv", "df_model_multi_annual.csv"))

df_model_bins <- df_model %>%
  mutate(map_bin = cut_width(pptMean, width = 0.4, boundary = 0)) %>%
  group_by(map_bin) %>%
  nest() %>%
  ungroup()

fit_gam <- function(data, spatial = FALSE, spatial_k = 12) {
  
  if (spatial) {
    model <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
                   s(pptCV, k = 8, bs = "ts") +
                   s(pptIntensity, k = 8, bs = "ts") +
                   s(pptGt95Pc, k = 8, bs = "ts") +
                   s(pptUgi, k = 8, bs = "ts") +
                   s(pptMeanDaySD, k = 8, bs = "ts") +
                   s(seasonLength, k = 8, bs = "ts") +
                   s(seasonOnsetSD, k = 8, bs = "ts") +
                   s(tMean, k = 8, bs = "ts") +
                   s(fire, k = 8, bs = "ts") +
                   s(sand, k = 8, bs = "ts") +
                   s(soilN, k = 8, bs = "ts") +
                   te(x, y, k = c(spatial_k, spatial_k), bs = "ts"),
                 data = data,
                 method = "REML",
                 family = Gamma(link = "log"),
                 select = TRUE,
                 samfrac = 0.1)
  } else {
    model <- bam(gpp ~ s(pptMean, k = 12, bs = "ts") +
                   s(pptCV, k = 8, bs = "ts") +
                   s(pptIntensity, k = 8, bs = "ts") +
                   s(pptGt95Pc, k = 8, bs = "ts") +
                   s(pptUgi, k = 8, bs = "ts") +
                   s(pptMeanDaySD, k = 8, bs = "ts") +
                   s(seasonLength, k = 8, bs = "ts") +
                   s(seasonOnsetSD, k = 8, bs = "ts") +
                   s(tMean, k = 8, bs = "ts") +
                   s(fire, k = 8, bs = "ts") +
                   s(sand, k = 8, bs = "ts") +
                   s(soilN, k = 8, bs = "ts"),
                 data = data,
                 method = "REML",
                 family = Gamma(link = "log"),
                 select = TRUE,
                 samfrac = 0.1)
  }
  
  model
}

# cl <- makeCluster(nc)
plan("multisession", workers = nc)

tic()
models_binned <- df_model_bins %>%
  mutate(mod1 = future_map(data, fit_gam, spatial = FALSE),
         mod2_k8 = future_map(data, fit_gam, spatial = TRUE, spatial_k = 8),
         mod2_k12 = future_map(data, fit_gam, spatial = TRUE, spatial_k = 12),
         mod2_k16 = future_map(data, fit_gam, spatial = TRUE, spatial_k = 16))
toc()
# stopCluster(cl)

write_rds(models_binned, here("results", "rds", "binned_models.rds"))


```


Visualise binned models:

```{r binned_models_viz}

models_binned <- read_rds(here("results", "rds", "binned_models.rds"))

# Drop final row (1600-2000 mm yr-1) - insufficient rows for model fitting
models_binned_cut <- models_binned[1:3,]

head(models_binned_cut)

# plot_gam <- function(gam) {
#   plot(gam, pages = 3, seWithMean = T, scheme = 2)
# }

# Set plot output directory

plot_dir <- here("results", "figures")

theme_set(theme_classic())

## Binned GAM partial effect plots

for (i in 1:nrow(models_binned_cut)) {
  
  # Extract smooth shapes as data frame using gratia
  
  mod1_smooths <- gratia::smooth_estimates(models_binned_cut$mod1[[i]], unconditional = TRUE) %>%
    mutate(Model = "Non-spatial",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000,
           pptCV = pptCV * 100)      # CHANGE THIS BACK WHEN I RERUN MODELS!
  mod2_k8_smooths <- gratia::smooth_estimates(models_binned_cut$mod2_k8[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - low\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000,
           pptCV = pptCV * 100) %>%
    select(-c(x,y))
  mod2_k12_smooths <- gratia::smooth_estimates(models_binned_cut$mod2_k12[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - medium\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000,
           pptCV = pptCV * 100) %>%
    select(-c(x,y))
  mod2_k16_smooths <- gratia::smooth_estimates(models_binned_cut$mod2_k16[[i]], unconditional = TRUE) %>%
    mutate(Model = "Spatial - high\ncomplexity",
           ci_low= exp(est - se * 1.96),
           ci_high = exp(est + se * 1.96),
           est = exp(est),
           pptMean = pptMean * 1000,
           pptCV = pptCV * 100) %>%
    select(-c(x,y))
  
  # Combine models into single data frame
  mod_smooths <- bind_rows(mod1_smooths,
                           mod2_k8_smooths,
                           mod2_k12_smooths,
                           mod2_k16_smooths) %>%
    filter(type == "TPRS (shrink)")
  
  # Read in data labels lookup table
  
  var_labels <- read_csv(here("data", "raw", "csv", "var_labels.csv"))
  
  var_list <- var_labels$vars
  
  plots <- list(length = length(var_list))
  
  # Generate plots
  
  plots <- map(var_list, function(var) {
    
    # Labels for each facet window
    # Create expressions for those with superscript/subscripts
    x_label <- var_labels$label[var_labels$vars == var]
    if (var_labels$parse[var_labels$vars == var]) {
      x_label <- parse(text = x_label)
    }
  
    # Plot data density plot
    data_df <- models_binned_cut$data[[i]] %>%
      select(x, y, all_of(var)) %>%
      rename(z = 3)
    
    density <- ggplot(data_df, aes(x = z)) +
      geom_density(colour = "grey70", fill = "grey90", alpha = 0.6) +
      theme_void()
    density
    
    # Plot model smooths
    smooth_df <- mod_smooths %>%
      select(smooth, type, est, se, ci_low, ci_high, Model,
               all_of(var)) %>%
      mutate(Model = factor(Model, levels = c("Non-spatial",
                                              "Spatial - low\ncomplexity",
                                              "Spatial - medium\ncomplexity",
                                              "Spatial - high\ncomplexity")),
             Model = ordered(Model)) %>%
      rename(var0 = 8) %>%
      drop_na()
    
    plot <- ggplot(smooth_df, aes(x = var0, y = est)) +
      geom_hline(yintercept = 1, colour = "grey80") +
      geom_ribbon(aes(ymin = ci_low, ymax = ci_high,
                      fill = Model), alpha = 0.7) +
      geom_line(aes(colour = Model, linetype = Model), alpha = 0.8, lwd = 0.6) +
      scale_colour_manual(values = c("#AAAAAA", "#bae4b3", "#31a354","#006d2c")) +
      scale_linetype_manual(values = c("longdash", "solid", "solid", "solid")) +
      scale_fill_manual(values = c("#DDDDDD", "#daffd3", "#51c384","#208d4c")) +
      guides(alpha = "none") +
      ylim(0, 2.5) +
      labs(x = x_label,
           y = "Partial effect")
    
      # if(var_labels$min_0[var_labels$vars == var] == 1) {
      #   plot <- plot + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
      #   density <- density + scale_x_continuous(limits = c(0, max(smooth_df$var0)))
      # }
    
    # Arrange into small multiples with patchwork package 
    density + plot +
        plot_layout(ncol = 1, nrow = 2, widths = 4, heights = c(1, 4))
    })
    
    # Create main GAM figure
    gam_plot <- ggarrange(plotlist = plots, ncol = 3, nrow = 4, common.legend = TRUE)
    
    # Save main GAM figure
    ggsave(here(plot_dir, paste0("binned_gam_plot_exp_", i, ".jpg")),
           plot = gam_plot,
           width = 20, height = 30, units = "cm", dpi = 300)
  
}

```

