---
title: "Changes in Africa's seasonal onset and cessation dates"
author: "Guy Lomax"
date: "2022-08-29"
output: html_notebook

---

This notebook processes data on wet season length, onset dates and cessation dates
and exports clean raster layers of wet season length, season length SD and onset
date SD for sub-Saharan Africa.

```{r setup, message = FALSE}

library(tidyverse)
library(terra)
library(sf)
library(tmap)
library(here)

```

Import data:

```{r load_data}

# Season length raster (single and double seasons)
season_length <- rast(here("data", "raw_data", "raster", "chirps",
                           "season_length_all.tif"))

# Seasonal rainfall annual dates and anomalies - single seasons
season_dates_s <- rast(here("data", "raw_data", "raster", "chirps",
                            "annual_dates_single.tif"))
seasonal_anomalies_s <- rast(here("data", "raw_data", "raster", "chirps",
                                  "season_deviation_single.tif"))

date_names <- c("onset_", "cessation_")
year_names <- as.character(rep(2001:2019, each = 2))

layer_names <- paste0(date_names, year_names)

names(season_dates_s) <- layer_names
names(seasonal_anomalies_s) <- layer_names

# Seasonal rainfall dates and anomalies - double seasons
season_dates_d <- rast(here("data", "raw_data", "raster", "chirps",
                            "annual_dates_double.tif"))
seasonal_anomalies_d <- rast(here("data", "raw_data", "raster", "chirps",
                                  "season_deviation_double.tif"))

date_names_d <- c("onset1_", "cessation1_", "onset2_", "cessation2_")
year_names_d <- as.character(rep(2001:2019, each = 4))

layer_names_d <- paste0(date_names_d, year_names_d)

names(season_dates_d) <- layer_names_d
names(seasonal_anomalies_d) <- layer_names_d

# Seasonality ratio

seasonality_ratio <- rast(here("data", "raw_data", "raster", "chirps",
                               "seasonality_ratio.tif")) %>%
  subset("amplitude_ratio")

# Africa country boundaries (Natural Earth dataset)
africa <- st_read(here("data", "raw_data", "vector", "natural_earth",
                       "ne_110m_admin_0_countries_fixed.shp")) %>%
  filter(CONTINENT == "Africa") %>%
  vect()

```

Mask and process single and double season rasters into single raster objects

``` {r pre-process}

# Function to mask to study area and separate single from double season pixels
mask_function <- function(rast, double = FALSE) {
  
  if(double == TRUE) {
    seasonality_mask <- seasonality_ratio >= 1
  } else {
    seasonality_mask <- seasonality_ratio < 1
  }

  mask(rast, seasonality_mask, maskvalues = c(0, NA))
}

# Mask single season rasters
season_dates_s <- mask_function(season_dates_s)
seasonal_anomalies_s <- mask_function(seasonal_anomalies_s)

# Mask double season rasters
season_dates_d <- mask_function(season_dates_d, TRUE)
seasonal_anomalies_d <- mask_function(seasonal_anomalies_d, TRUE)


```

Explore data and identify issues/errors:


```{r eda, eval = FALSE}

# summary(seasonal_anomalies_s)
# summary(seasonal_anomalies_d)

# Single season onset dates
onset_anomalies_s <- seasonal_anomalies_s[[grepl("onset", names(seasonal_anomalies_s))]]

a <- sum(values(onset_anomalies_s) < -60 | values(onset_anomalies_s) > 120, na.rm = TRUE)
b <- sum(!is.na(values(onset_anomalies_s)))
print("Single season onset")
a/b

# Double season onset dates
onset_anomalies_d <- seasonal_anomalies_d[[grepl("onset", names(seasonal_anomalies_d))]]


c <- sum(values(onset_anomalies_d) < -45 | values(onset_anomalies_d) > 60, na.rm = TRUE)
d <- sum(!is.na(values(onset_anomalies_d)))
print("Double season onset")
c/d

# Single season cessation dates
cessation_anomalies_s <- seasonal_anomalies_s[[grepl("cessation", names(seasonal_anomalies_s))]]

# summary(cessation_anomalies)
e <- sum(values(cessation_anomalies_s) < -120 | values(cessation_anomalies_s) > 120, na.rm = TRUE)
f <- sum(!is.na(values(cessation_anomalies_s)))
print("Single season cessation")
e/f

# Double season cessation dates
cessation_anomalies_d <- seasonal_anomalies_d[[grepl("cessation", names(seasonal_anomalies_d))]]

# summary(cessation_anomalies)
g <- sum(values(cessation_anomalies_d) < -60 | values(cessation_anomalies_d) > 45, na.rm = TRUE)
h <- sum(!is.na(values(cessation_anomalies_d)))
print("Double season cessation")
g/h

# Identify pixels that have values outside the expected range

onset_error_pixels <- sum(onset_anomalies_s < -60 | onset_anomalies_s > 120)
cessation_error_pixels <- sum(cessation_anomalies_s < -120 | cessation_anomalies_s > 120)
onset_error_pixels_d <- sum(onset_anomalies_d < -45 | onset_anomalies_d > 60)
cessation_error_pixels_d <- sum(cessation_anomalies_d < -60 | cessation_anomalies_d > 45)

all_onset_errors <- merge(onset_error_pixels, onset_error_pixels_d)
all_cessation_errors <- merge(cessation_error_pixels, cessation_error_pixels_d)

plot(all_onset_errors)
lines(africa)
plot(all_cessation_errors)
lines(africa)

```

Remove pixels/seasons where onset/cessation dates have not been properly
identified:

```{r date_anomaly_pre_processing}

# Create a mask for removing seasons/years where one date is undetermined
anomaly_mask_s <- (onset_anomalies_s >= -60 &
                          onset_anomalies_s <= 120) & 
                       (cessation_anomalies_s >= -120 &
                          cessation_anomalies_s <= 120)

anomaly_mask_d <- (onset_anomalies_d >= -45 &
                          onset_anomalies_d <= 60) & 
                       (cessation_anomalies_d >= -60 &
                          cessation_anomalies_d <= 45)

# Fully mask pixels where more than 20% of years/seasons include NAs
lack_of_data_s <- 
  (sum(anomaly_mask_s) > nlyr(anomaly_mask_s) * 0.8)
lack_of_data_d <- 
  (sum(anomaly_mask_d) > nlyr(anomaly_mask_d) * 0.8)

lack_of_data <- merge(lack_of_data_s, lack_of_data_d)

# Apply masks

NA_mask <- function(rast, double = FALSE) {
  if(double == TRUE) {
    anomaly_mask <- anomaly_mask_d
  } else {
    anomaly_mask <- anomaly_mask_s
  }
  
  rast %>%
    mask(anomaly_mask, maskvalues = c(0, NA)) %>%
    mask(lack_of_data, maskvalues = c(0, NA))
}

onset_anomalies_s_masked <- NA_mask(onset_anomalies_s)
cessation_anomalies_s_masked <- NA_mask(cessation_anomalies_s)
onset_anomalies_d_masked <- NA_mask(onset_anomalies_d, double = TRUE)
cessation_anomalies_d_masked <- NA_mask(cessation_anomalies_d, double = TRUE)

# Apply the same mask to the raw season onset/cessation layers

onset_dates_s <- season_dates_s[[grepl("onset", names(season_dates_s))]] %>%
  NA_mask()
cessation_dates_s <- season_dates_s[[grepl("cessation", names(season_dates_s))]] %>%
  NA_mask()

onset_dates_d <- season_dates_d[[grepl("onset", names(season_dates_d))]] %>%
  NA_mask(double = TRUE)
cessation_dates_d <- season_dates_d[[grepl("cessation", names(season_dates_d))]] %>%
  NA_mask(double = TRUE)

# summary(onset_anomalies_s_masked)
# summary(cessation_anomalies_s_masked)
# summary(onset_anomalies_d_masked)
# summary(cessation_anomalies_d_masked)

```

After masking, calculate key long-term metrics and combine single and double
into a single layer:

```{r seasonal_metrics}

# Calculate total season lengths and merge
annual_length_s <- cessation_dates_s - onset_dates_s
season_length_d <- (cessation_dates_d - onset_dates_d)

# For double, sum individual season lengths for each year
annual_length_d <- season_length_d[[seq(1, nlyr(season_length_d)-1, 2)]] +
  season_length_d[[seq(2, nlyr(season_length_d), 2)]]

annual_length_all <- merge(annual_length_s, annual_length_d)

# Standard deviation of onset, cessation and length
# Need to define a function to calculate SD that tolerates NA values while
# returning NA for fully masked pixels

season_sd <- function(x) {
  if(all(is.na(x))) {
    NA
  } else {
    sd(x, na.rm = TRUE)
  }
}

# Apply to relevant spatRasters
season_length_sd <- app(annual_length_all, fun = season_sd)

# Calculate onset sd
season_onset_sd_s <- app(onset_anomalies_s_masked, fun = season_sd)
season_onset_sd_d <- app(onset_anomalies_d_masked, fun = season_sd)
season_onset_sd <- merge(season_onset_sd_s, season_onset_sd_d)

```

Export data:

```{r export}

# Export as data layers for use in modelling
writeRaster(season_length_sd, "processed_data/raster/seasonality/season_length_sd.tif",
            overwrite = TRUE)
writeRaster(season_onset_sd, "processed_data/raster/seasonality/season_onset_sd.tif",
            overwrite = TRUE)

```

